---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

```{r, include=FALSE}
library(shapeshiftr)
library(data.table)
```

# shapeshiftr

<!-- badges: start -->
[![R-CMD-check](https://github.com/lkhenayfis/shapeshiftr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lkhenayfis/shapeshiftr/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/lkhenayfis/shapeshiftr/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/lkhenayfis/shapeshiftr/actions/workflows/test-coverage.yaml)
[![codecov](https://codecov.io/gh/lkhenayfis/shapeshiftr/graph/badge.svg?token=JW6LQ7TO2U)](https://codecov.io/gh/lkhenayfis/shapeshiftr)
<!-- badges: end -->

Package `shapeshiftr` aims at providing a simple and efficient interface for manipulation and 
transformation of data for training forecasting models. This is effectively done through "slicing"
operations, which transform standard tabular time series data into data.frames usable for model
fitting/forecasting, as well as diverse transforms (such as wavelets).

## Installation

You can install the development version of shapeshiftr from
[GitHub](https://github.com/lkhenayfis/shapeshiftr) with:

``` r
# install.packages("devtools")
devtools::install_github("lkhenayfis/shapeshiftr")
```

## Main functionalities

### Pipes

The `shapeshiftr` package provides a flexible pipe system for defining data transformations through
closure generators. Pipes allow you to define transformation workflows that can be parsed and
evaluated on data.

#### Simple transformation pipes

A pipe transformation is defined by a closure generator - a function that returns another function.
Here's a simple example that computes summary statistics:

```{r}
# Define a closure generator for summary statistics
gen_closure_summary <- function(...) function(x) summary(x)

# Define a pipe that applies this transformation to mtcars
raw_pipe <- list(
    on = "mtcars",
    transforms = list(
        list(fun = "gen_closure_summary")
    )
)

# Parse the pipe to generate the closure
parsed_pipe <- parse_single_pipe(raw_pipe)

# Evaluate the pipe on the data
result <- eval_single_pipe(parsed_pipe, env = list(mtcars = mtcars))
result
```

#### Pipes with computed parameters

More sophisticated pipes can perform computations and use the results as parameters in the returned
closure. This example shows a scaling transformation that centers data to zero mean and scales to
unit variance:

```{r}
# Define a closure generator that computes scaling parameters
gen_closure_scale <- function(x, ...) {
    # Compute mean and sd from the training data x (happens once during parsing)
    means <- colMeans(x, na.rm = TRUE)
    sds <- apply(x, 2, sd, na.rm = TRUE)
    
    # Return a closure that uses these pre-computed parameters
    function(x) {
        as.data.frame(scale(x, center = means, scale = sds))
    }
}

# Define a pipe that scales the data
raw_pipe <- list(
    on = "mtcars",
    transforms = list(
        list(fun = "gen_closure_scale")
    )
)

# Parse and evaluate
parsed_pipe <- parse_single_pipe(raw_pipe, env = list(mtcars = mtcars), enclos = environment())
scaled_data <- eval_single_pipe(parsed_pipe, env = list(mtcars = mtcars), enclos = environment())
head(scaled_data)
```

#### Chaining multiple transformations

Pipes can chain multiple transformations together. Each transformation
receives the output of the previous one:

```{r}
# Define a filter closure generator
gen_closure_filter <- function(by, value, ...) {
    function(x) x[x[[by]] == value, ]
}

# Combine filtering and summary
raw_pipe <- list(
    on = "mtcars",
    transforms = list(
        list(fun = "gen_closure_filter", by = "cyl", value = 6),
        list(fun = "gen_closure_summary")
    )
)

# Parse and evaluate
parsed_pipe <- parse_single_pipe(raw_pipe, env = list(mtcars = mtcars), enclos = environment())
result <- eval_single_pipe(parsed_pipe, env = list(mtcars = mtcars), enclos = environment())
result
```

### Slicing

This is a basic example that demonstrates how `shapeshiftr` is meant to aid in data manipulation in
it's canonical use case.

Suppose there are two datasets:

* `obs`, containing observed data of explanatory variables "X1", "X2" and target variable "Y"

```{r}
obs  <- simple_dt_date[, .(date, X1, X2, Y)]
print(head(obs))
```

* `pred`, containing predicted variables of the regressors "X1" and "X2" (i.e. numerical weather forecasts)

```{r}
pred <- keyed_dt_date[, .(date, target_date, X1, X2)]
print(head(pred, 10))
```

The goal is to forecast "Y" two days ahead using 3 lags of "X1" and "X2", plus their forecasted
values two days ahead. To build this dataset, we use

```{r}
lag_X <- slice(obs, "date", variables = c("X1", "X2"), L = -3:-1,
    names = c("lag_X1", "lag_X2"))
lead_X <- slice(pred, "date", "target_date", variables = c("X1", "X2"), L = 0:2,
    names = c("lead_X1", "lead_X2"))
target <- slice(obs, "date", variables = "Y", L = 2)

X <- merge(lag_X, lead_X)
training_data <- merge(X, target)

dt <- as.data.table(training_data)
```

The resulting data.table is

```{r}
print(head(dt, 10))
```

## Full Example: Training and Inference Consistency

One of the most powerful features of the pipe system is its ability to maintain transformation
consistency between training and inference. When training a forecasting model, you often need to
apply transformations (like scaling) to your data. During inference, you must apply the *exact same*
transformations using the *same parameters* computed from the training data.

This example demonstrates how to:

1. Create pipes that scale data then perform slicing
2. Parse and evaluate pipes on training data (computing scaling parameters)
3. Re-use the same parsed pipes on test data (applying the same scaling parameters)

```{r}
# Split the data into training (first half) and test (second half)
n <- nrow(obs)
train_obs <- obs[1:floor(n / 2)]
test_obs <- obs[(floor(n / 2) + 1):n]
```

Now we define closure generators for scaling and slicing:

```{r}
# Closure generator that computes scaling parameters and returns a function
# that applies those parameters
gen_closure_scale_cols <- function(x, cols, ...) {
    # Compute scaling parameters from x (training data) - happens ONCE during parsing
    means <- colMeans(x[, cols, with = FALSE], na.rm = TRUE)
    sds <- sapply(x[, cols, with = FALSE], sd, na.rm = TRUE)

    # Return a closure that uses these pre-computed, fixed parameters
    function(x) {
        result <- copy(x)
        for (col in cols) {
            result[[col]] <- (x[[col]] - means[col]) / sds[col]
        }
        result
    }
}

# Closure generator for slicing
gen_closure_slice_simple <- function(x, index_var, variables, L, ...) {
    # The returned closure performs slicing
    # x is the training data but we don't use it here - slicing doesn't need "training"
    function(x) {
        result <- slice(x, index_var, variables = variables, L = L)
        as.data.table(result)
    }
}
```

Define the pipe that first scales, then slices:

```{r}
# Create a pipe that scales X1, X2, Y, then creates lagged features
pipe_def <- list(
    on = "data",
    transforms = list(
        # First: scale the data
        list(
            fun = "gen_closure_scale_cols",
            cols = c("X1", "X2", "Y")
        ),
        # Then: create lagged features
        list(
            fun = "gen_closure_slice_simple",
            index_var = "date",
            variables = c("X1", "X2", "Y"),
            L = -2:0
        )
    )
)
```

**Step 1: Parse and evaluate on training data**

```{r}
# Parse the pipe using TRAINING data - this computes scaling parameters from training data
parsed_pipe_train <- parse_single_pipe(pipe_def,
    env = list(data = train_obs),
    enclos = environment())

# Evaluate on training data
train_result <- eval_single_pipe(parsed_pipe_train,
    env = list(data = train_obs),
    enclos = environment())

cat("Training data after scaling and slicing:\n")
print(head(train_result, 5))
print(tail(train_result, 5))
```

**Step 2: Evaluate the SAME parsed pipe on test data**

The key insight: we use `parsed_pipe_train` (which contains scaling parameters from training) on test data:

```{r}
# Evaluate the SAME parsed pipe on test data
# This applies the scaling parameters learned from TRAINING data
test_result <- eval_single_pipe(parsed_pipe_train,
    env = list(data = test_obs),
    enclos = environment())

cat("\nTest data after applying TRAINING scaling and slicing:\n")
print(head(test_result, 5))
print(tail(test_result, 5))
```

This workflow demonstrates the core value proposition of the pipe system: **transformation
consistency**. By parsing pipes on training data and reusing them on test/inference data, you
ensure that all transformations use the same parameters, which is critical for model performance
and avoiding data leakage.
