---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

```{r, include=FALSE}
library(shapeshiftr)
library(data.table)
```

# shapeshiftr

<!-- badges: start -->
<!-- badges: end -->

Package `shapeshiftr` aims at providing a simple and efficient interface for manipulation and 
transformation of data for training forecasting models. This is effectively done through "slicing"
operations, which transform standard tabular time series data into data.frames usable for model
fitting/forecasting, as well as diverse transforms (such as wavelets).

## Installation

You can install the development version of shapeshiftr from
[GitHub](https://github.com/lkhenayfis/shapeshiftr) with:

``` r
# install.packages("devtools")
devtools::install_github("lkhenayfis/shapeshiftr")
```

## Example

This is a basic example that demonstrates how `shapeshiftr` is meant to aid in data manipulation in
it's canonical use case.

Suppose there are two datasets:

* `obs`, containing observed data of explanatory variables "X1", "X2" and target variable "Y"

```{r}
obs  <- simple_dt_date[, .(date, X1, X2, Y)]
print(obs)
```

* `pred`, containing predicted variables of the regressors "X1" and "X2" (i.e. numerical weather forecasts)

```{r}
pred <- keyed_dt_date[, .(date, target_date, X1, X2)]
print(head(pred, 15))
```

The goal is to forecast "Y" two days ahead using 3 lags of "X1" and "X2", plus their forecasted
values two days ahead. To build this dataset, we use

```{r}
lag_X <- slice(obs, "date", variables = c("X1", "X2"), L = -3:-1,
    names = c("lag_X1", "lag_X2"))
lead_X <- slice(pred, "date", "target_date", variables = c("X1", "X2"), L = 0:2,
    names = c("lead_X1", "lead_X2"))
target <- slice(obs, "date", variables = "Y", L = 2)

X <- merge(lag_X, lead_X)
training_data <- merge(X, target)

dt <- as.data.table(training_data)
print(head(dt, 10))
```
