---
title: "Transformation Consistency with Pipes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transformation Consistency with Pipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

```{r libraries, message = FALSE}
library(shapeshiftr)
library(data.table)

data(simple_dt_date)
```

## Introduction

Forecasting model development faces a critical challenge: **ensuring transformation consistency between training, testing, and production data**.

Consider this common scenario:

1. You scale your training data to mean=0, sd=1 using **training set statistics**
2. You fit a model on the scaled training data
3. You need to evaluate on test data — but how do you scale it?

**WRONG approach**: Compute mean and sd from TEST data and scale  
**Result**: Data leakage! Test set statistics influenced the transformation  
**Consequence**: Invalid model evaluation, overly optimistic performance estimates

**RIGHT approach**: Use the SAME mean and sd from TRAINING data to scale test  
**Result**: True held-out evaluation  
**Challenge**: How to "remember" training parameters and apply them consistently?

shapeshiftr's **pipes system** solves this through closure-based transformation generators that:

- Compute parameters ONCE from training data (parse phase)
- Apply those parameters MANY TIMES to train, test, and production data (eval phase)
- Guarantee transformation consistency across all datasets

### The Data Leakage Problem

Let's see this problem in action with a concrete example:

```{r data_leakage_example}
set.seed(123)
train_idx <- sample(seq_len(nrow(simple_dt_date)), 15)
train_data <- simple_dt_date[train_idx, .(date, X1, X2, Y)]
test_data <- simple_dt_date[-train_idx, .(date, X1, X2, Y)]

# Training data - X1 statistics:
cat(sprintf("  Mean: %g\n  SD: %g\n\n",
        mean(train_data$X1), sd(train_data$X1)))

# Test data - X1 statistics (DIFFERENT):
cat(sprintf("  Mean: %g\n  SD: %g\n\n",
        mean(test_data$X1), sd(test_data$X1)))

# WRONG: Scale test data using TEST statistics (DATA LEAKAGE!)
test_mean_WRONG <- mean(test_data$X1)
test_sd_WRONG <- sd(test_data$X1)
test_scaled_WRONG <- (test_data$X1 - test_mean_WRONG) / test_sd_WRONG

# WRONG approach - Test data scaled with TEST stats:
# Problem: Test statistics influenced the transformation!
cat(sprintf("  Mean: %g (should be ~0)\n  SD: %g (should be ~1)\n\n",
        round(mean(test_scaled_WRONG), 10),
        round(sd(test_scaled_WRONG), 10)))

# RIGHT: Scale test data using TRAINING statistics
train_mean <- mean(train_data$X1)
train_sd <- sd(train_data$X1)
test_scaled_RIGHT <- (test_data$X1 - train_mean) / train_sd

# RIGHT approach - Test data scaled with TRAINING stats:
# Success: True held-out evaluation!
cat(sprintf("  Mean: %g (NOT 0, as expected)\n  SD: %g (NOT 1, as expected)\n",
        round(mean(test_scaled_RIGHT), 3),
        round(sd(test_scaled_RIGHT), 3)))
```

**Key Insight**: The WRONG approach gives perfect 0/1 statistics because it used test data to compute them. The RIGHT approach gives different statistics, proving it's truly held-out.

### The Pipes Solution

shapeshiftr's pipes system automates the RIGHT approach through **closure generators**:

```{r solution_preview, eval = FALSE}
gen_scale <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols])
    train_sds <- apply(x[, ..cols], 2, sd)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}

parsed_pipe <- parse_single_pipe(
    list(on = "data", transforms = list(list(fun = "gen_scale", cols = "X1"))),
    env = list(data = train_data)
)

train_scaled <- eval_single_pipe(parsed_pipe, env = list(data = train_data))
test_scaled <- eval_single_pipe(parsed_pipe, env = list(data = test_data))
```

Don't worry if this looks complex — Sections 2-4 will build your understanding step by step.

## What You'll Learn

This vignette teaches transformation consistency through eight progressive sections:

1. **Introduction** (this section): The train-test consistency problem
2. **Pipe Fundamentals**: Closure generator architecture and parse-eval workflow
3. **Basic Usage Patterns**: Simple stateless and trainable transformations
4. **Recipe-Based Transformations**: Five transformation generators (scaling, encoding, log, winsorization, Box-Cox)
5. **Custom Closure Generators**: Writing your own domain-specific transformations
6. **Advanced Patterns**: Chaining transformations and composition strategies
7. **Integration with Slicing**: Combining pipes with temporal slicing for complete workflows
8. **Complete Forecasting Pipeline**: End-to-end example with categorical encoding and slicing

**Estimated reading time**: 20-25 minutes

**Prerequisites**: 
- Basic R knowledge (functions, lists, data.tables)
- Familiarity with forecasting concepts (train/test split, feature engineering)
- Optional: Read `vignette("slicing")` first for temporal slicing context

Let's begin by understanding the closure architecture.

## Pipe Fundamentals

The pipes system separates transformation into two distinct phases:

- **Parse phase**: Compute parameters from training data, generate closures
- **Eval phase**: Apply closures to current data (train, test, or production)

This separation is the key to train-test consistency.

### Closure Generator Architecture

A **closure generator** is a function that returns a function. The returned function (the closure) "remembers" variables from its parent environment.

Here's the anatomy of a trainable closure generator:

```{r closure_anatomy}
gen_standardize <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**What happens**:

1. Generator body executes ONCE during parse phase
2. Computes `train_means` and `train_sds` from training data `x`
3. Returns a closure that captures these computed parameters
4. Closure executes MANY TIMES during eval phase
5. Each eval uses the SAME captured parameters

**Key architectural points**:

- NO `parse` or `eval` arguments in the generator
- NO `if (parse) {...} else {...}` branching logic
- Separation is architectural (parse vs eval functions), not conditional
- Generator accepts `x` (training data) and `...` (for interface consistency)
- Closure only accepts `x` (current data to transform)

### Parse-Eval Workflow

The workflow uses two core functions: `parse_single_pipe()` and `eval_single_pipe()`.

**Parse Phase** (happens ONCE with training data):

```{r parse_workflow}
train <- simple_dt_date[1:15]
data_env <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env)
```

What `parse_single_pipe()` does:

1. Looks up `"data"` in `data_env` (finds training data)
2. Calls `gen_standardize(cols = c("X1", "X2"), x = data_env$data, ...)`
3. Generator computes training statistics and returns closure
4. Stores closure in `parsed$transforms[[1]]`

**Eval Phase** (happens MANY TIMES with different data):

```{r eval_workflow}
train_result <- eval_single_pipe(parsed, env = data_env)

# Training data transformation:
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

test <- simple_dt_date[16:20]
data_env$data <- test

test_result <- eval_single_pipe(parsed, env = data_env)

# Test data transformation (uses TRAINING parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))
```

What `eval_single_pipe()` does:

1. Looks up `"data"` in current `data_env` (could be train or test)
2. Calls stored closure with current data: `closure(x = data_env$data)`
3. Closure applies captured training parameters to current data
4. Returns transformed data

**Critical insight**: The SAME `parsed` pipe is used for both train and test, guaranteeing identical parameters.

### Named List Data Swapping

The recommended pattern uses named lists to enable easy data swapping:

```{r named_list_pattern}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_list_train <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_list_train)

train_transformed <- eval_single_pipe(parsed, env = data_list_train)

data_list_test <- list(data = test)

test_transformed <- eval_single_pipe(parsed, env = data_list_test)
```

**Why this pattern**:

- **Parse once**: Parameters computed from training data
- **Eval many**: Apply to train, test, validation, production
- **Easy swapping**: Change `env` parameter without modifying pipe structure
- **Clear intent**: `on = "data"` references list element, not global variable

### Train-Test Consistency Demonstration

Let's verify the complete workflow ensures consistency:

```{r consistency_demo}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env)

closure <- parsed$transforms[[1]]
closure_env <- environment(closure)
captured_means <- get("train_means", closure_env)
captured_sds <- get("train_sds", closure_env)

# Captured training parameters:
cat(sprintf("Training mean X1: %g\nTraining sd X1: %g\n\n",
        captured_means["X1"], captured_sds["X1"]))

train_result <- eval_single_pipe(parsed, env = data_env)

# Training transformation (mean ≈ 0, sd ≈ 1):
cat(sprintf("X1 - Mean: %g, SD: %g\n\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

data_env$data <- test
test_result <- eval_single_pipe(parsed, env = data_env)

# Test transformation (uses TRAINING parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))

manual_verify <- (test$X1 - captured_means["X1"]) / captured_sds["X1"]
cat(sprintf("Manual verification match: %s\n",
        all.equal(manual_verify, test_result$X1)))
```

**Proof of consistency**: Test transformation uses captured training parameters, not test statistics.

## Basic Usage Patterns

Closure generators fall into two categories: **stateless** and **trainable**.

### Stateless Transformations

Stateless generators apply fixed operations without learning from training data.

**Example: Log transformation**

```{r stateless_example}
gen_log <- function(cols, offset = 1, ...) {
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- log(x_copy[[col]] + offset)
        }
        x_copy
    }
}

data_env <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log", cols = "Y", offset = 1)
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env)
result <- eval_single_pipe(parsed, env = data_env)

# Original Y (first 5):
print(head(data_env$data$Y, 5))

# Logged Y (first 5):
cat("\n")
print(head(result$Y, 5))

# Verification:
cat(sprintf("\nExpected: %g\nGot: %g\n",
        log(data_env$data$Y[1] + 1), result$Y[1]))
```

**Pattern**: No computation from `x` in generator body, all parameters are arguments.

### Trainable Transformations

Trainable generators learn parameters from training data.

**Example: Standardization**

```{r trainable_example}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env)

train_result <- eval_single_pipe(parsed, env = data_env)

# Training result:
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

data_env$data <- test
test_result <- eval_single_pipe(parsed, env = data_env)

# Test result (uses training parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))
```

**Pattern**: Generator body computes parameters from `x`, closure captures them.

### Environment Management

Both `parse_single_pipe()` and `eval_single_pipe()` accept `env` and `enclos` parameters:

- **`env`**: Where to look up data references (e.g., `"data"` in `on = "data"`)
  - Can be a named list, data.frame, or environment
  - Controls DATA scope
  
- **`enclos`**: Where to look up function references (e.g., `"gen_standardize"`)
  - Must be an environment
  - Controls FUNCTION scope

**Example: Custom function environment**

```{r env_management}
custom_env <- new.env()
assign("gen_standardize", gen_standardize, envir = custom_env)

data_list <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_list, enclos = custom_env)
result <- eval_single_pipe(parsed, env = data_list, enclos = custom_env)

cat(sprintf("X1 mean: %g\n", round(mean(result$X1), 10)))
```

**Use cases**:
- Package development (isolated namespaces)
- Testing (controlled environments)
- Deployment (function versioning)

### Error Handling Basics

Generators should validate inputs:

```{r error_handling, error = TRUE}
gen_standardize_safe <- function(cols, x, ...) {
    if (!is.data.table(x) && !is.data.frame(x)) {
        stop("x must be a data.table or data.frame")
    }

    missing_cols <- setdiff(cols, names(x))
    if (length(missing_cols) > 0) {
        stop("Columns not found: ", paste(missing_cols, collapse = ", "))
    }

    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}

data_env <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize_safe", cols = c("X1", "MISSING"))
    )
)

try(parse_single_pipe(raw_pipe, env = data_env))
```

**Best practices**:
- Validate inputs in generator body (fails fast during parse)
- Check column existence
- Handle NA values appropriately
- Provide clear error messages

## Recipe-Based Transformations

> **shapeshiftr as Infrastructure Framework**
>
> The transformation generators demonstrated in this section showcase the pipes framework's capabilities. However, **shapeshiftr is infrastructure, not a pre-built transformation toolkit**. These examples live in the vignette to teach the framework — they are not exported functions.
>
> For production forecasting workflows with extensive transformation needs, consider:
>
> - **recipes package**: Comprehensive pre-built transformation library for general ML
> - **Future companion package**: Curated shapeshiftr-compatible generator library (planned)
> - **Custom generators**: Write domain-specific transformations using patterns shown here (Section 5)
>
> shapeshiftr's value is providing the **closure-based parse-eval infrastructure** that ensures train-test consistency. You bring the transformations.

### What This Section Demonstrates

This section shows 5 common transformation types implemented as closure generators:

1. **Scaling**: Standardization with training mean/sd
2. **Categorical Encoding**: One-hot encoding with training factor levels
3. **Mathematical Transform**: Log transformation (stateless)
4. **Outlier Handling**: Winsorization at training quantiles
5. **Variance Stabilization**: Box-Cox power transform

Each example demonstrates the CORRECT closure pattern and train-test consistency.

### Generator 1: Standardization (gen_scale_cols)

**Purpose**: Standardize numeric columns to mean=0, sd=1 using training statistics

**When to Use**: 
- Models sensitive to scale (neural networks, SVM, regularized regression)
- Features with different units or magnitudes
- Before PCA or clustering

**Pattern**: Trainable (learns mean and sd from training data)

**Implementation**:

```{r gen_scale_cols}
gen_scale_cols <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**Demo: Train-Test Consistency**

```{r demo_scale_cols}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_cols", cols = c("X1", "X2", "X3"))
    )
)

parsed_scale <- parse_single_pipe(pipe_scale, env = data_env)

train_scaled <- eval_single_pipe(parsed_scale, env = data_env)

# Training data (scaled):
cat(sprintf("X1 - Mean: %g, SD: %g\nX2 - Mean: %g, SD: %g\n\n",
        round(mean(train_scaled$X1), 6), round(sd(train_scaled$X1), 6),
        round(mean(train_scaled$X2), 6), round(sd(train_scaled$X2), 6)))

data_env$data <- test

test_scaled <- eval_single_pipe(parsed_scale, env = data_env)

# Test data (uses training mean/sd):
# Different from 0/1 - correctly uses training stats
cat(sprintf("X1 - Mean: %g, SD: %g\nX2 - Mean: %g, SD: %g\n\n",
        round(mean(test_scaled$X1), 4), round(sd(test_scaled$X1), 4),
        round(mean(test_scaled$X2), 4), round(sd(test_scaled$X2), 4)))

cat("Train-test consistency: test uses training parameters\n")
```

**Key Insight**: Test data scaled with training mean/sd prevents data leakage

### Generator 2: One-Hot Encoding (gen_one_hot)

**Purpose**: Convert categorical variables to binary dummy variables using training factor levels

**When to Use**:
- Linear models requiring numeric inputs
- Tree-based models (optional, they can handle factors directly)
- When you need explicit control over reference level

**Pattern**: Trainable (learns factor levels from training data)

**Implementation**:

```{r gen_one_hot}
gen_one_hot <- function(col, x, drop_first = TRUE, ...) {
    if (!is.factor(x[[col]])) {
        x[[col]] <- as.factor(x[[col]])
    }
    train_levels <- levels(x[[col]])

    if (drop_first) {
        encode_levels <- train_levels[-1]
    } else {
        encode_levels <- train_levels
    }

    function(x) {
        x_copy <- copy(x)

        for (lev in encode_levels) {
            col_name <- paste0(col, "_", lev)
            x_copy[[col_name]] <- as.integer(x_copy[[col]] == lev)
        }

        x_copy[[col]] <- NULL
        x_copy
    }
}
```

**Demo: Categorical Encoding**

```{r demo_one_hot}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

simple_dt_date[, day_type := as.factor(day_type)]

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

pipe_onehot <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_one_hot", col = "day_type", drop_first = TRUE)
    )
)

parsed_onehot <- parse_single_pipe(pipe_onehot, env = data_env)

train_encoded <- eval_single_pipe(parsed_onehot, env = data_env)

# Original day_type (training):
print(head(train$day_type, 8))

# Encoded columns (training):
cat(sprintf("\nColumns created: %s\n",
        paste(setdiff(names(train_encoded), names(train)), collapse = ", ")))
print(head(train_encoded[, .(date, day_type_weekend)], 8))

data_env$data <- test

test_encoded <- eval_single_pipe(parsed_onehot, env = data_env)

# Test data encoded with training levels:
cat("\n")
print(head(test_encoded[, .(date, day_type_weekend)], 5))

cat("\nOne-hot encoding uses training levels on test data\n")
```

**Key Insight**: Test data encoded using training factor levels prevents unseen level errors

### Generator 3: Log Transformation (gen_log_transform)

**Purpose**: Apply log transformation with optional offset

**When to Use**:
- Right-skewed distributions
- Multiplicative relationships → additive
- Variance stabilization for count data
- Before modeling with linear assumptions

**Pattern**: Stateless (no training needed, offset is fixed)

**Implementation**:

```{r gen_log_transform}
gen_log_transform <- function(cols, offset = 1, base = exp(1), ...) {
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            if (base == exp(1)) {
                x_copy[[col]] <- log(x_copy[[col]] + offset)
            } else if (base == 10) {
                x_copy[[col]] <- log10(x_copy[[col]] + offset)
            } else if (base == 2) {
                x_copy[[col]] <- log2(x_copy[[col]] + offset)
            } else {
                x_copy[[col]] <- log(x_copy[[col]] + offset) / log(base)
            }
        }
        x_copy
    }
}
```

**Demo: Log Transformation**

```{r demo_log_transform}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

data_env <- list(data = simple_dt_date[1:15])

pipe_log <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log_transform", cols = "Y", offset = 1)
    )
)

parsed_log <- parse_single_pipe(pipe_log, env = data_env)

train_logged <- eval_single_pipe(parsed_log, env = data_env)

# Original Y (training, first 5):
print(head(data_env$data$Y, 5))

# Logged Y (training, first 5):
cat("\n")
print(head(train_logged$Y, 5))

# Verification: log(Y + 1)
cat(sprintf("\nExpected: %g\nGot: %g\n\n",
        log(data_env$data$Y[1] + 1), train_logged$Y[1]))

data_env$data <- simple_dt_date[16:20]
test_logged <- eval_single_pipe(parsed_log, env = data_env)

# Test data also log-transformed:
print(head(test_logged$Y, 5))

cat("\nStateless: Same transformation applied to all data\n")
```

**Key Insight**: Stateless transformations apply fixed operations uniformly

### Generator 4: Winsorization (gen_winsorize)

**Purpose**: Clip extreme values at training data quantiles

**When to Use**:
- Reduce outlier influence without removal
- Robust transformation before scaling
- Preserve sample size while limiting extremes
- When outliers are measurement errors, not true extremes

**Pattern**: Trainable (learns quantile bounds from training data)

**Implementation**:

```{r gen_winsorize}
gen_winsorize <- function(cols, x, lower = 0.05, upper = 0.95, ...) {
    bounds <- lapply(cols, function(col) {
        quantile(x[[col]], c(lower, upper), na.rm = TRUE, names = FALSE)
    })
    names(bounds) <- cols

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            lower_bound <- bounds[[col]][1]
            upper_bound <- bounds[[col]][2]
            x_copy[[col]] <- pmin(pmax(x_copy[[col]], lower_bound), upper_bound)
        }
        x_copy
    }
}
```

**Demo: Outlier Clipping**

```{r demo_winsorize}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

pipe_winsor <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_winsorize", cols = c("X1", "X2"),
            lower = 0.1, upper = 0.9)
    )
)

parsed_winsor <- parse_single_pipe(pipe_winsor, env = data_env)

train_winsor <- eval_single_pipe(parsed_winsor, env = data_env)

# Training X1:
bounds_10_90 <- quantile(train$X1, c(0.1, 0.9))
cat(sprintf("Original range: %g to %g\nWinsorized range: %g to %g\n",
        min(train$X1), max(train$X1),
        min(train_winsor$X1), max(train_winsor$X1)))
cat(sprintf("10th/90th percentiles (bounds): %g, %g\n\n",
        bounds_10_90[1], bounds_10_90[2]))

data_env$data <- test

test_winsor <- eval_single_pipe(parsed_winsor, env = data_env)

# Test X1:
# Uses training bounds, not test quantiles
cat(sprintf("Original range: %g to %g\nWinsorized range: %g to %g\n\n",
        min(test$X1), max(test$X1),
        min(test_winsor$X1), max(test_winsor$X1)))

cat("Winsorization clips test data at training quantiles\n")
```

**Key Insight**: Test outliers clipped using training quantiles maintains consistency

### Generator 5: Box-Cox Transform (gen_boxcox)

**Purpose**: Power transformation for variance stabilization and normality

**When to Use**:
- Heteroscedastic errors (variance increases with mean)
- Non-normal distributions
- Before linear regression with normality assumptions
- When sqrt or log transforms insufficient

**Pattern**: Trainable (estimates lambda from training data)

**Note**: This is a simplified implementation using fixed lambda=0.5 (square root). Production code would estimate optimal lambda via maximum likelihood.

**Implementation**:

```{r gen_boxcox}
gen_boxcox <- function(cols, x, ...) {
    lambdas <- setNames(rep(0.5, length(cols)), cols)

    cat(sprintf("Box-Cox lambdas (simplified to 0.5 for demo): %s\n\n",
            paste(names(lambdas), lambdas, sep = "=", collapse = ", ")))

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            lambda <- lambdas[col]
            if (lambda == 0) {
                x_copy[[col]] <- log(x_copy[[col]])
            } else {
                x_copy[[col]] <- (x_copy[[col]]^lambda - 1) / lambda
            }
        }
        x_copy
    }
}
```

**Demo: Box-Cox Transform**

```{r demo_boxcox}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env <- list(data = train)

pipe_boxcox <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_boxcox", cols = "X3")
    )
)

parsed_boxcox <- parse_single_pipe(pipe_boxcox, env = data_env)

train_bc <- eval_single_pipe(parsed_boxcox, env = data_env)

# Training X3:
cat(sprintf("Original - Mean: %g, SD: %g\nBox-Cox - Mean: %g, SD: %g\n\n",
        round(mean(train$X3), 3), round(sd(train$X3), 3),
        round(mean(train_bc$X3), 3), round(sd(train_bc$X3), 3)))

data_env$data <- test

test_bc <- eval_single_pipe(parsed_boxcox, env = data_env)

# Test X3 (uses training lambda=0.5):
cat(sprintf("Original - Mean: %g\nBox-Cox - Mean: %g\n\n",
        round(mean(test$X3), 3), round(mean(test_bc$X3), 3)))

cat("Box-Cox applies training lambda to test data\n")
```

**Key Insight**: Test data transformed with training lambda ensures consistency

### Transformation Summary

| Generator | Type | Purpose | Learns From Training |
|-----------|------|---------|---------------------|
| `gen_scale_cols` | Trainable | Standardization (mean=0, sd=1) | Mean, SD |
| `gen_one_hot` | Trainable | Categorical encoding | Factor levels |
| `gen_log_transform` | Stateless | Log transformation | None (fixed offset) |
| `gen_winsorize` | Trainable | Outlier clipping | Quantile bounds |
| `gen_boxcox` | Trainable | Variance stabilization | Lambda parameter |

**Common Pattern**: All trainable generators:
1. Compute parameters from training data in generator body
2. Return closure that captures those parameters
3. Closure applies parameters to current data
4. Test data uses training parameters → train-test consistency

**Code Location**: These generators are demonstration code living in this vignette. They are **not exported functions**. For production use, write your own generators (Section 5) or use a dedicated transformation library like recipes.

### Key Takeaways

- **5 transformation types demonstrated**: scaling, one-hot encoding, log, winsorization, Box-Cox
- **Trainable generators** (4 of 5): Learn parameters from training data during parse
- **Stateless generators** (1 of 5): Apply fixed transformations without training
- **CORRECT closure pattern**: All generators use parameter computation in generator body, application in closure
- **Train-test consistency**: All demos show test data using training parameters
- **Framework positioning**: shapeshiftr is infrastructure, these are learning examples (not package exports)
- **Categorical handling**: `gen_one_hot` enables mixed-type feature engineering
- **Data leakage prevention**: Architecture makes it impossible to compute test statistics

**Production Usage**:
- Use these patterns to write custom generators for your domain (Section 5)
- Consider recipes package for extensive pre-built transformations
- Combine multiple transformations with closure composition (Section 6)

Next, we'll learn how to **write custom closure generators** for domain-specific transformations (Section 5).
