---
title: "Transformation Consistency with Pipes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transformation Consistency with Pipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

```{r libraries, message = FALSE}
library(shapeshiftr)
library(data.table)

data(simple_dt_date)
```

## Introduction

Forecasting model development faces a critical challenge: **ensuring transformation consistency between training, testing, and production data**.

Consider this common scenario:

1. You scale your training data to mean=0, sd=1 using **training set statistics**
2. You fit a model on the scaled training data
3. You need to evaluate on test data — but how do you scale it?

**WRONG approach**: Compute mean and sd from TEST data and scale  
**Result**: Data leakage! Test set statistics influenced the transformation  
**Consequence**: Invalid model evaluation, overly optimistic performance estimates

**RIGHT approach**: Use the SAME mean and sd from TRAINING data to scale test  
**Result**: True held-out evaluation  
**Challenge**: How to "remember" training parameters and apply them consistently?

shapeshiftr's **pipes system** solves this through closure-based transformation generators that:

- Compute parameters ONCE from training data (parse phase)
- Apply those parameters MANY TIMES to train, test, and production data (eval phase)
- Guarantee transformation consistency across all datasets

### The Data Leakage Problem

Let's see this problem in action with a concrete example:

```{r data_leakage_example}
set.seed(123)
train_idx <- sample(seq_len(nrow(simple_dt_date)), 15)
train_data <- simple_dt_date[train_idx, .(date, X1, X2, Y)]
test_data <- simple_dt_date[-train_idx, .(date, X1, X2, Y)]

# Training data - X1 statistics:
cat(sprintf("  Mean: %g\n  SD: %g\n\n",
        mean(train_data$X1), sd(train_data$X1)))

# Test data - X1 statistics (DIFFERENT):
cat(sprintf("  Mean: %g\n  SD: %g\n\n",
        mean(test_data$X1), sd(test_data$X1)))

# WRONG: Scale test data using TEST statistics (DATA LEAKAGE!)
test_mean_WRONG <- mean(test_data$X1)
test_sd_WRONG <- sd(test_data$X1)
test_scaled_WRONG <- (test_data$X1 - test_mean_WRONG) / test_sd_WRONG

# WRONG approach - Test data scaled with TEST stats:
# Problem: Test statistics influenced the transformation!
cat(sprintf("  Mean: %g (should be ~0)\n  SD: %g (should be ~1)\n\n",
        round(mean(test_scaled_WRONG), 10),
        round(sd(test_scaled_WRONG), 10)))

# RIGHT: Scale test data using TRAINING statistics
train_mean <- mean(train_data$X1)
train_sd <- sd(train_data$X1)
test_scaled_RIGHT <- (test_data$X1 - train_mean) / train_sd

# RIGHT approach - Test data scaled with TRAINING stats:
# Success: True held-out evaluation!
cat(sprintf("  Mean: %g (NOT 0, as expected)\n  SD: %g (NOT 1, as expected)\n",
        round(mean(test_scaled_RIGHT), 3),
        round(sd(test_scaled_RIGHT), 3)))
```

**Key Insight**: The WRONG approach gives perfect 0/1 statistics because it used test data to compute them. The RIGHT approach gives different statistics, proving it's truly held-out.

### The Pipes Solution

shapeshiftr's pipes system automates the RIGHT approach through **closure generators**:

```{r solution_preview, eval = FALSE}
gen_scale <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols])
    train_sds <- apply(x[, ..cols], 2, sd)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}

parsed_pipe <- parse_single_pipe(
    list(on = "data", transforms = list(list(fun = "gen_scale", cols = "X1"))),
    env = list(data = train_data)
)

train_scaled <- forward_single_pipe(parsed_pipe, env = list(data = train_data))
test_scaled <- forward_single_pipe(parsed_pipe, env = list(data = test_data))
```

Don't worry if this looks complex — Sections 2-4 will build your understanding step by step.

## What You'll Learn

This vignette teaches transformation consistency through eight progressive sections:

1. **Introduction** (this section): The train-test consistency problem
2. **Pipe Fundamentals**: Closure generator architecture and parse-eval workflow
3. **Basic Usage Patterns**: Simple stateless and trainable transformations and combining multiple pipes
4. **Recipe-Based Transformations**: Five transformation generators (scaling, encoding, log, winsorization, Box-Cox)
5. **Custom Closure Generators**: Writing your own domain-specific transformations
6. **Advanced Patterns**: Chaining transformations, composition strategies
7. **Integration with Slicing**: Combining pipes with temporal slicing for complete workflows
8. **Complete Forecasting Pipeline**: End-to-end example with categorical encoding and slicing

**Estimated reading time**: 20-25 minutes

**Prerequisites**: 
- Basic R knowledge (functions, lists, data.tables)
- Familiarity with forecasting concepts (train/test split, feature engineering)
- Optional: Read `vignette("slicing")` first for temporal slicing context

Let's begin by understanding the closure architecture.

## Pipe Fundamentals

The pipes system separates transformation into two distinct phases:

- **Parse phase**: Compute parameters from training data, generate closures
- **Eval phase**: Apply closures to current data (train, test, or production)

This separation is the key to train-test consistency.

### Closure Generator Architecture

A **closure generator** is a function that returns a function. The returned function (the closure) "remembers" variables from its parent environment.

Here's the anatomy of a trainable closure generator:

```{r closure_anatomy}
gen_standardize <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**What happens**:

1. Generator body executes ONCE during parse phase
2. Computes `train_means` and `train_sds` from training data `x`
3. Returns a closure that captures these computed parameters
4. Closure executes MANY TIMES during eval phase
5. Each eval uses the SAME captured parameters

**Key architectural points**:

- NO `parse` or `eval` arguments in the generator
- NO `if (parse) {...} else {...}` branching logic
- Separation is architectural (parse vs eval functions), not conditional
- Generator accepts `x` (training data) and `...` (for interface consistency)
- Closure only accepts `x` (current data to transform)

### Parse-Eval Workflow

The workflow uses two core functions: `parse_single_pipe()` and `forward_single_pipe()`.

**Parse Phase** (happens ONCE with training data):

```{r parse_workflow}
train <- simple_dt_date[1:15]
data_env_train <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env_train)
```

What `parse_single_pipe()` does:

1. Looks up `"data"` in `data_env_train` (finds training data)
2. Calls `gen_standardize(cols = c("X1", "X2"), x = data_env_train$data, ...)`
3. Generator computes training statistics and returns closure
4. Stores closure in `parsed$transforms[[1]]`

**Eval Phase** (happens MANY TIMES with different data):

```{r eval_workflow}
train_result <- forward_single_pipe(parsed, env = data_env_train)

# Training data transformation:
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

test <- simple_dt_date[16:20]
data_env_test <- list(data = test)

test_result <- forward_single_pipe(parsed, env = data_env_test)

# Test data transformation (uses TRAINING parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))
```

What `forward_single_pipe()` does:

1. Looks up `"data"` in current `data_env_test` (could be train or test)
2. Calls stored closure with current data: `closure(x = data_env_test$data)`
3. Closure applies captured training parameters to current data
4. Returns transformed data

**Critical insight**: The SAME `parsed` pipe is used for both train and test, guaranteeing identical parameters.

### Named List Data Swapping

The recommended pattern uses named lists to enable easy data swapping:

```{r named_list_pattern}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_list_train <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_list_train)

train_transformed <- forward_single_pipe(parsed, env = data_list_train)

data_list_test <- list(data = test)

test_transformed <- forward_single_pipe(parsed, env = data_list_test)
```

**Why this pattern**:

- **Parse once**: Parameters computed from training data
- **Eval many**: Apply to train, test, validation, production
- **Easy swapping**: Change `env` parameter without modifying pipe structure
- **Clear intent**: `on = "data"` references list element, not global variable

### Train-Test Consistency Demonstration

Let's verify the complete workflow ensures consistency:

```{r consistency_demo}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env_train)

closure <- parsed$transforms[[1]]$forward
closure_env <- environment(closure)
captured_means <- get("train_means", closure_env)
captured_sds <- get("train_sds", closure_env)

# Captured training parameters:
cat(sprintf("Training mean X1: %g\nTraining sd X1: %g\n\n",
        captured_means["X1"], captured_sds["X1"]))

train_result <- forward_single_pipe(parsed, env = data_env_train)

# Training transformation (mean ≈ 0, sd ≈ 1):
cat(sprintf("X1 - Mean: %g, SD: %g\n\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

data_env_test <- list(data = test)
test_result <- forward_single_pipe(parsed, env = data_env_test)

# Test transformation (uses TRAINING parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))

manual_verify <- (test$X1 - captured_means["X1"]) / captured_sds["X1"]
cat(sprintf("Manual verification match: %s\n",
        all.equal(manual_verify, test_result$X1)))
```

**Proof of consistency**: Test transformation uses captured training parameters, not test statistics.

## Basic Usage Patterns

Closure generators fall into two categories: **stateless** and **trainable**.

### Stateless Transformations

Stateless generators apply fixed operations without learning from training data.

**Example: Log transformation**

```{r stateless_example}
gen_log <- function(cols, offset = 1, ...) {
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- log(x_copy[[col]] + offset)
        }
        x_copy
    }
}

data_env <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log", cols = "Y", offset = 1)
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env)
result <- forward_single_pipe(parsed, env = data_env)

# Original Y (first 5):
print(head(data_env$data$Y, 5))

# Logged Y (first 5):
cat("\n")
print(head(result$Y, 5))

# Verification:
cat(sprintf("\nExpected: %g\nGot: %g\n",
        log(data_env$data$Y[1] + 1), result$Y[1]))
```

**Pattern**: No computation from `x` in generator body, all parameters are arguments.

### Trainable Transformations

Trainable generators learn parameters from training data.

**Example: Standardization**

```{r trainable_example}
train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_env_train)

train_result <- forward_single_pipe(parsed, env = data_env_train)

# Training result:
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(train_result$X1), 10),
        round(sd(train_result$X1), 6)))

data_env_test <- list(data = test)
test_result <- forward_single_pipe(parsed, env = data_env_test)

# Test result (uses training parameters):
cat(sprintf("X1 - Mean: %g, SD: %g\n",
        round(mean(test_result$X1), 4),
        round(sd(test_result$X1), 4)))
```

**Pattern**: Generator body computes parameters from `x`, closure captures them.

### Combining Multiple Pipes

So far `raw_pipe` has always defined transformations regarding a single variable. In practice you
may want to define multiple pipes for different variables or transformation sequences.

```{r combining_pipes}
raw_pipes <- list(
    list(
        on = "data_first_half",
        transforms = list(
            list(fun = "gen_standardize", cols = c("X1", "X2"))
        )
    ),
    list(
        on = "data_second_half",
        transforms = list(
            list(fun = "gen_standardize", cols = c("X1", "X2"))
        )
    )
)

data_list <- list(
    data_first_half = simple_dt_date[1:10],
    data_second_half = simple_dt_date[11:20]
)

parsed <- parse_pipes(raw_pipes, env = data_list)

result <- forward_pipes(parsed, env = data_list)

# combining by rbind, for example
comb_fun <- function(x1, x2) rbind(x1, x2)
combine_pipes(result, comb_fun)
```

In this simple example we combined the results manually using `combine_pipes()` with rbind. More
complex combination strategies can be implemented as needed. In fact, one could wrap everything in
pipes as well through a more heterdox use of the pipes described later in Advanced Patterns.

### Environment Management

Both `parse_single_pipe()` and `forward_single_pipe()` accept `env` and `enclos` parameters:

- **`env`**: Where to look up data references (e.g., `"data"` in `on = "data"`)
  - Can be a named list, data.frame, or environment
  - Controls DATA scope
  
- **`enclos`**: Where to look up function references (e.g., `"gen_standardize"`)
  - Must be an environment
  - Controls FUNCTION scope

**Example: Custom function environment**

```{r env_management}
custom_env <- new.env()
assign("gen_standardize", gen_standardize, envir = custom_env)

data_list <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(raw_pipe, env = data_list, enclos = custom_env)
result <- forward_single_pipe(parsed, env = data_list, enclos = custom_env)

cat(sprintf("X1 mean: %g\n", round(mean(result$X1), 10)))
```

**Use cases**:
- Package development (isolated namespaces)
- Testing (controlled environments)
- Deployment (function versioning)

### Error Handling Basics

Generators should validate inputs:

```{r error_handling, error = TRUE}
gen_standardize_safe <- function(cols, x, ...) {
    if (!is.data.table(x) && !is.data.frame(x)) {
        stop("x must be a data.table or data.frame")
    }

    missing_cols <- setdiff(cols, names(x))
    if (length(missing_cols) > 0) {
        stop("Columns not found: ", paste(missing_cols, collapse = ", "))
    }

    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}

data_env <- list(data = simple_dt_date[1:15])

raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_standardize_safe", cols = c("X1", "MISSING"))
    )
)

try(parse_single_pipe(raw_pipe, env = data_env))
```

**Best practices**:
- Validate inputs in generator body (fails fast during parse)
- Check column existence
- Handle NA values appropriately
- Provide clear error messages

## Reversible Transformations

Many forecasting workflows require **bidirectional transformations**: transforming features or targets before modeling, then inverting those transformations on predictions to return to original units. This section introduces the reversible transformation pattern that makes this workflow seamless and train-test consistent.

shapeshiftr provides `forward_single_pipe()` and `backward_single_pipe()` functions that automatically:
1. Extract the appropriate forward or backward closures from parsed pipes
2. Apply them via the standard `forward_single_pipe()` mechanism
3. Reverse the transformation order for backward direction (ensuring mathematical correctness)

### Introduction: Why Reversibility Matters

In forecasting pipelines, you frequently encounter scenarios requiring transformation inversion:

**Target transformation workflow**:
1. Transform target variable (e.g., log transform for variance stabilization)
2. Fit model on transformed target
3. Generate predictions in transformed space
4. **Invert transformation** to get forecasts in original units

**Feature transformation workflow**:
1. Transform features (e.g., PCA for dimensionality reduction)
2. Fit model on transformed features
3. Generate predictions
4. **Invert feature transformation** for interpretation or feature importance

**Critical challenge**: The inversion MUST use the SAME parameters learned during forward transformation from training data. Using different parameters breaks train-test consistency and invalidates your pipeline.

**Example**: If you standardize a target variable using training mean=100 and sd=15, you must unstandardize predictions using those SAME values (not test statistics). Otherwise, your forecasts will be on the wrong scale.

**The reversible generator pattern** solves this by returning both `forward` and `backward` closures that share the same captured training parameters via closure environment. This architectural guarantee makes parameter inconsistency impossible.

### Creating Reversible Generators

A reversible generator returns a list containing two closures: `forward` (applies transformation) and `backward` (inverts transformation). Both closures share the same parent environment containing training parameters.

**Single closure (not reversible)**:
```{r reversible_intro_single, eval = FALSE}
gen_scale <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols])
    train_sds <- apply(x[, ..cols], 2, sd)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**Paired closures (reversible)**:
```{r reversible_intro_paired}
gen_scale_reversible <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    forward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }

    backward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] * train_sds[col]) + train_means[col]
        }
        x_copy
    }

    list(forward = forward, backward = backward)
}
```

**Key architectural differences**:
- Single closure: Returns one function (only forward transformation)
- Paired closures: Returns `list(forward = ..., backward = ...)`
- Both closures capture `train_means` and `train_sds` from shared parent environment
- Forward applies `(x - mean) / sd`, backward applies `(x * sd) + mean`
- No duplication of training parameters (single source of truth)

**Environment sharing demonstration**:
```{r reversible_env_sharing}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]

data_env_train <- list(data = train)

pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_reversible", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(pipe_scale, env = data_env_train)

forward_closure <- parsed$transforms[[1]]$forward
backward_closure <- parsed$transforms[[1]]$backward

identical(environment(forward_closure), environment(backward_closure))

forward_env <- environment(forward_closure)
captured_means <- get("train_means", forward_env)
captured_sds <- get("train_sds", forward_env)

cat(sprintf("Captured means: X1=%g, X2=%g\n", captured_means["X1"], captured_means["X2"]))
cat(sprintf("Captured sds: X1=%g, X2=%g\n", captured_sds["X1"], captured_sds["X2"]))
```

**Proof of shared environment**: Both closures reference the SAME environment object, guaranteeing parameter consistency.

### Forward vs Backward Modes

The forward and backward closures represent two **directions** of transformation, not two **phases** of the pipeline. Both are used during the evaluation phase.

shapeshiftr provides `forward_single_pipe()` and `backward_single_pipe()` to apply transformations in both directions while automatically handling closure extraction and order reversal.

**Direction semantics**:
- **Forward**: Original units → Transformed units (for model input)
- **Backward**: Transformed units → Original units (for interpretation/output)

**Workflow integration**:

```{r reversible_workflow_demo, eval = FALSE}
parsed <- parse_single_pipe(pipe_def, env = list(data = train))

train_transformed <- forward_single_pipe(parsed, env = list(data = train))

model <- fit_model(Y ~ X1 + X2, data = train_transformed)

test_transformed <- forward_single_pipe(parsed, env = list(data = test))

predictions_transformed <- predict(model, newdata = test_transformed)

predictions_original <- backward_single_pipe(
    parsed,
    env = list(data = data.table(Y = predictions_transformed))
)
```

**Order reversal in backward transformation**:

When chaining multiple transformations, backward direction automatically reverses the order:

```{r reversible_order, eval = FALSE}
pipe_chain <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log_reversible", cols = "Y", offset = 1),
        list(fun = "gen_scale_reversible", cols = "Y")
    )
)

parsed_chain <- parse_single_pipe(pipe_chain, env = list(data = train))

train_forward <- forward_single_pipe(parsed_chain, env = list(data = train))

train_reconstructed <- backward_single_pipe(parsed_chain, env = list(data = train_forward))
```

**Why reverse order**: Transformations compose mathematically as functions. If forward is `g(f(x))`, backward is `f^(-1)(g^(-1)(x))`, not `g^(-1)(f^(-1)(x))`. The `backward_single_pipe()` function handles this reversal automatically.

### Roundtrip Validation

A critical property of reversible transformations is **roundtrip consistency**: applying forward then backward should reconstruct the original data (within numerical precision for exact transformations, within tolerance for approximate transformations).

**Testing exact reversibility**:

```{r reversible_roundtrip_exact}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15, .(date, X1, X2)]

data_env <- list(data = train)

pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_reversible", cols = c("X1", "X2"))
    )
)

parsed <- parse_single_pipe(pipe_scale, env = data_env)

train_scaled <- forward_single_pipe(parsed, env = data_env)

train_reconstructed <- backward_single_pipe(parsed, env = list(data = train_scaled))

max_error_X1 <- max(abs(train$X1 - train_reconstructed$X1))
max_error_X2 <- max(abs(train$X2 - train_reconstructed$X2))

cat(sprintf("Max reconstruction error X1: %g\n", max_error_X1))
cat(sprintf("Max reconstruction error X2: %g\n", max_error_X2))
cat(sprintf("Exact reversibility (error < 1e-10): %s\n",
        max_error_X1 < 1e-10 && max_error_X2 < 1e-10))
```

**Testing approximate reversibility**:

Some transformations are inherently lossy (e.g., PCA with dimensionality reduction). For these, roundtrip validation verifies error is within acceptable bounds.

```{r reversible_roundtrip_approx}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

gen_pca <- function(cols, n_components = 2, x, ...) {
    pca_model <- prcomp(x[, ..cols], center = TRUE, scale. = TRUE)

    rotation_matrix <- pca_model$rotation[, seq_len(n_components), drop = FALSE]
    center <- pca_model$center
    scale <- pca_model$scale

    forward <- function(x) {
        x_copy <- copy(x)

        x_scaled <- scale(x[, ..cols], center = center, scale = scale)
        pc_scores <- x_scaled %*% rotation_matrix

        for (i in seq_len(n_components)) {
            pc_name <- paste0("PC", i)
            x_copy[[pc_name]] <- pc_scores[, i]
        }

        for (col in cols) {
            x_copy[[col]] <- NULL
        }

        x_copy
    }

    backward <- function(x) {
        x_copy <- copy(x)

        pc_cols <- paste0("PC", seq_len(n_components))
        pc_matrix <- as.matrix(x[, ..pc_cols])

        x_reconstructed <- pc_matrix %*% t(rotation_matrix)
        x_reconstructed <- sweep(x_reconstructed, 2, scale, "*")
        x_reconstructed <- sweep(x_reconstructed, 2, center, "+")

        for (i in seq_along(cols)) {
            x_copy[[cols[i]]] <- x_reconstructed[, i]
        }

        for (pc_col in pc_cols) {
            x_copy[[pc_col]] <- NULL
        }

        x_copy
    }

    list(forward = forward, backward = backward)
}

train <- simple_dt_date[1:15, .(date, X1, X2, X3)]

data_env <- list(data = train)

pipe_pca <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_pca", cols = c("X1", "X2", "X3"), n_components = 2)
    )
)

parsed <- parse_single_pipe(pipe_pca, env = data_env)

train_pca <- forward_single_pipe(parsed, env = data_env)

train_reconstructed <- backward_single_pipe(parsed, env = list(data = train_pca))

mean_error_X1 <- mean(abs(train$X1 - train_reconstructed$X1))
mean_error_X2 <- mean(abs(train$X2 - train_reconstructed$X2))
mean_error_X3 <- mean(abs(train$X3 - train_reconstructed$X3))

cat(sprintf("Mean reconstruction error X1: %g\n", mean_error_X1))
cat(sprintf("Mean reconstruction error X2: %g\n", mean_error_X2))
cat(sprintf("Mean reconstruction error X3: %g\n", mean_error_X3))

relative_error <- mean(c(mean_error_X1, mean_error_X2, mean_error_X3)) /
    mean(c(sd(train$X1), sd(train$X2), sd(train$X3)))

cat(sprintf("Relative reconstruction error: %g%%\n", relative_error * 100))
cat(sprintf("Approximate reversibility acceptable (error < 10%%): %s\n",
        relative_error < 0.1))
```

**Validation best practices**:
- **Exact transformations**: Verify roundtrip error < 1e-10 (floating point precision)
- **Approximate transformations**: Define acceptable relative error threshold (e.g., 10% of feature standard deviation)
- **Test on training data**: Roundtrip should work on data used to learn parameters
- **Test on test data**: Roundtrip should also work on held-out data
- **Automate in tests**: Include roundtrip checks in unit tests for custom generators

### Train-Test Consistency in Reversible Transformations

The reversible pattern guarantees train-test consistency in BOTH forward and backward directions by sharing the same captured parameters.

**Forward consistency** (already covered in Section 2-3):

```{r reversible_consistency_forward}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15, .(date, X1)]
test <- simple_dt_date[16:20, .(date, X1)]

data_env_train <- list(data = train)

pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_reversible", cols = "X1")
    )
)

parsed <- parse_single_pipe(pipe_scale, env = data_env_train)

train_scaled <- forward_single_pipe(parsed, env = data_env_train)

cat(sprintf("Train X1: mean=%g, sd=%g\n",
        round(mean(train_scaled$X1), 10), round(sd(train_scaled$X1), 6)))

data_env_test <- list(data = test)

test_scaled <- forward_single_pipe(parsed, env = data_env_test)

cat(sprintf("Test X1: mean=%g, sd=%g\n",
        round(mean(test_scaled$X1), 4), round(sd(test_scaled$X1), 4)))

cat("\nForward: Test uses training mean/sd (not 0/1)\n")
```

**Backward consistency** (the new guarantee):

```{r reversible_consistency_backward}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15, .(date, Y)]
test <- simple_dt_date[16:20, .(date, Y)]

gen_log_reversible <- function(cols, offset = 1, base = exp(1), ...) {
    forward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            if (base == exp(1)) {
                x_copy[[col]] <- log(x_copy[[col]] + offset)
            } else {
                x_copy[[col]] <- log(x_copy[[col]] + offset) / log(base)
            }
        }
        x_copy
    }

    backward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            if (base == exp(1)) {
                x_copy[[col]] <- exp(x_copy[[col]]) - offset
            } else {
                x_copy[[col]] <- base^x_copy[[col]] - offset
            }
        }
        x_copy
    }

    list(forward = forward, backward = backward)
}

data_env_train <- list(data = train)

pipe_log <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log_reversible", cols = "Y", offset = 5)
    )
)

parsed <- parse_single_pipe(pipe_log, env = data_env_train)

train_logged <- forward_single_pipe(parsed, env = data_env_train)

predictions_logged <- data.table(Y = train_logged$Y[1:5])

predictions_original <- backward_single_pipe(
    parsed,
    env = list(data = predictions_logged)
)

cat("Logged predictions:\n")
print(predictions_logged)

cat("\nBack-transformed predictions (offset=5 from training):\n")
print(predictions_original)

cat("\nOriginal training Y (first 5):\n")
print(train[1:5, .(Y)])

cat(sprintf("\nBackward used training offset=%d consistently\n", 5))
```

**Key insight**: The backward closure uses the SAME offset parameter (5) that was specified during parse phase, not a different value. This consistency applies to all captured parameters (means, sds, rotation matrices, offsets, etc.).

**Production scenario**:

Imagine you fit a model on log-transformed targets:

1. **Parse** with training data, offset=5 (learns this parameter)
2. **Forward transform** training targets using offset=5
3. Fit model on transformed targets
4. **Forward transform** test targets using offset=5 (consistency!)
5. Generate predictions in log space
6. **Backward transform** predictions using offset=5 (consistency!)
7. Predictions now in original units with correct scale

**Without reversible pattern**: You might accidentally use different offsets in forward (training) vs backward (prediction), producing forecasts on the wrong scale. The reversible pattern makes this error architecturally impossible.

### Best Practices for Reversible Transformations

**1. Always test roundtrip consistency during development**:

```{r reversible_bestpractice_roundtrip, eval = FALSE}
test_that("gen_my_transform roundtrip is exact", {
    x <- data.table(col1 = rnorm(100))
    gen <- gen_my_transform(cols = "col1", x = x)

    x_transformed <- gen$forward(x)
    x_reconstructed <- gen$backward(x_transformed)

    expect_equal(x_reconstructed$col1, x$col1, tolerance = 1e-10)
})
```

**2. Document reversibility type (exact vs approximate)**:

```{r reversible_bestpractice_doc, eval = FALSE}
#' @description
#'     Applies standardization transformation with exact reversibility.
#'     Roundtrip reconstruction: backward(forward(x)) == x within floating
#'     point precision.
```

**3. Name parameters consistently**:

Always use `forward` and `backward` (not `transform`/`inverse`, `encode`/`decode`, etc.) for consistency with shapeshiftr conventions.

**4. Handle edge cases in both directions**:

```{r reversible_bestpractice_edge, eval = FALSE}
backward <- function(x) {
    x_copy <- copy(x)
    for (col in cols) {
        if (any(train_sds[col] == 0)) {
            warning(sprintf("Column %s had sd=0 in training, cannot unstandardize", col))
            next
        }
        x_copy[[col]] <- (x_copy[[col]] * train_sds[col]) + train_means[col]
    }
    x_copy
}
```

**5. Return both closures even if only forward is currently used**:

Future workflow changes may require backward transformation. Providing both closures from the start prevents refactoring.

**6. Verify shared environment in tests**:

```{r reversible_bestpractice_env, eval = FALSE}
test_that("forward and backward share environment", {
    gen <- gen_scale_reversible(cols = "X", x = data.table(X = 1:10))

    expect_identical(
        environment(gen$forward),
        environment(gen$backward)
    )
})
```

### Common Patterns for Reversible Generators

**Pattern 1: Exact linear transformation**
- Examples: Scaling, centering, affine transformations
- Reversibility: Exact
- Implementation: Compute inverse operation algebraically

**Pattern 2: Exact nonlinear transformation**
- Examples: Log/exp, Box-Cox (with fixed lambda)
- Reversibility: Exact
- Implementation: Apply inverse function

**Pattern 3: Approximate dimensionality reduction**
- Examples: PCA, SVD, autoencoders
- Reversibility: Approximate (information loss)
- Implementation: Project to lower dimension, then reconstruct (won't be exact)

**Pattern 4: Stateless reversible**
- Examples: Fixed offset log transform, deterministic encoding
- Reversibility: Exact
- Implementation: No training needed, both directions use fixed parameters

**Pattern 5: Trainable reversible**
- Examples: Standardization, PCA, quantile mapping
- Reversibility: Exact or approximate depending on method
- Implementation: Learn parameters from training data, use in both directions

**Framework positioning**: These patterns demonstrate how to structure reversible generators within shapeshiftr's closure-based architecture. Use them as templates for domain-specific bidirectional transformations.

### Complete Roundtrip Workflow: Data Exploration

This example demonstrates using PCA for dimensionality reduction with backward transformation to interpret results in original feature space. PCA is inherently lossy when reducing dimensions (3 → 2), so backward reconstruction is approximate.

**Workflow**:
1. Apply PCA to reduce 3 features to 2 principal components
2. Parse and forward transform training data
3. Identify interesting data points in PC space (extreme PC1 values)
4. Backward transform selected points to interpret in original units
5. Verify roundtrip accuracy (expect information loss from dimensionality reduction)

```{r data_exploration_roundtrip}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15, .(date, X1, X2, X3)]

data_env_train <- list(data = train)

pipe_pca <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_pca", cols = c("X1", "X2", "X3"), n_components = 2)
    )
)

parsed_pca <- parse_single_pipe(pipe_pca, env = data_env_train)

train_pca <- forward_single_pipe(parsed_pca, env = data_env_train)

cat("Forward transformation (3D -> 2D PCA):\n")
print(train_pca[1:5, .(date, PC1, PC2)])

extreme_pc1_idx <- which.max(abs(train_pca$PC1))
cat(sprintf("\nIdentified extreme PC1 at row %d: PC1=%g, PC2=%g\n",
        extreme_pc1_idx, train_pca$PC1[extreme_pc1_idx],
        train_pca$PC2[extreme_pc1_idx]))

extreme_point <- train_pca[extreme_pc1_idx]

reconstructed <- backward_single_pipe(parsed_pca, env = list(data = extreme_point))

cat("\nBackward transformation (2D -> 3D reconstruction):\n")
cat(sprintf("Original: X1=%g, X2=%g, X3=%g\n",
        train$X1[extreme_pc1_idx], train$X2[extreme_pc1_idx],
        train$X3[extreme_pc1_idx]))
cat(sprintf("Reconstructed: X1=%g, X2=%g, X3=%g\n",
        reconstructed$X1, reconstructed$X2, reconstructed$X3))

error_X1 <- abs(train$X1[extreme_pc1_idx] - reconstructed$X1)
error_X2 <- abs(train$X2[extreme_pc1_idx] - reconstructed$X2)
error_X3 <- abs(train$X3[extreme_pc1_idx] - reconstructed$X3)
cat(sprintf("Reconstruction errors: X1=%g, X2=%g, X3=%g\n",
        error_X1, error_X2, error_X3))

subset_idx <- 1:5
train_subset_pca <- forward_single_pipe(
    parsed_pca,
    env = list(data = train[subset_idx])
)

train_subset_reconstructed <- backward_single_pipe(
    parsed_pca,
    env = list(data = train_subset_pca)
)

mean_error_X1 <- mean(abs(train[subset_idx]$X1 - train_subset_reconstructed$X1))
mean_error_X2 <- mean(abs(train[subset_idx]$X2 - train_subset_reconstructed$X2))
mean_error_X3 <- mean(abs(train[subset_idx]$X3 - train_subset_reconstructed$X3))

cat(sprintf("\nRoundtrip verification on 5 points:\n"))
cat(sprintf("Mean errors: X1=%g, X2=%g, X3=%g\n",
        mean_error_X1, mean_error_X2, mean_error_X3))
cat("Approximate reversibility: Information loss expected (3D -> 2D -> 3D)\n")
```

**Key insights**:
- PCA reduces 3 features to 2 components (lossy transformation)
- Backward transformation enables interpretation of PC space findings in original units
- Roundtrip reconstruction shows information loss from dimensionality reduction
- Both forward and backward use training PCA parameters (rotation matrix, center, scale)
- Workflow demonstrates: parse → forward → identify patterns → backward → interpret

This pattern applies to any approximate reversible transformation where you need to move between transformed and original spaces for analysis.

## Recipe-Based Transformations

> **shapeshiftr as Infrastructure Framework**
>
> The transformation generators demonstrated in this section showcase the pipes framework's capabilities. However, **shapeshiftr is infrastructure, not a pre-built transformation toolkit**. These examples live in the vignette to teach the framework — they are not exported functions.
>
> For production forecasting workflows with extensive transformation needs, consider:
>
> - **recipes package**: Comprehensive pre-built transformation library for general ML
> - **Future companion package**: Curated shapeshiftr-compatible generator library (planned)
> - **Custom generators**: Write domain-specific transformations using patterns shown here (Section 5)
>
> shapeshiftr's value is providing the **closure-based parse-eval infrastructure** that ensures train-test consistency. You bring the transformations.

### What This Section Demonstrates

This section shows 5 common transformation types implemented as closure generators:

1. **Scaling**: Standardization with training mean/sd
2. **Categorical Encoding**: One-hot encoding with training factor levels
3. **Mathematical Transform**: Log transformation (stateless)
4. **Outlier Handling**: Winsorization at training quantiles
5. **Variance Stabilization**: Box-Cox power transform

Each example demonstrates the CORRECT closure pattern and train-test consistency.

### Generator 1: Standardization (gen_scale_cols)

**Purpose**: Standardize numeric columns to mean=0, sd=1 using training statistics

**When to Use**: 
- Models sensitive to scale (neural networks, SVM, regularized regression)
- Features with different units or magnitudes
- Before PCA or clustering

**Pattern**: Trainable (learns mean and sd from training data)

**Implementation**:

```{r gen_scale_cols}
gen_scale_cols <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**Reversible version** (for target scaling workflows):

```{r gen_scale_cols_reversible}
gen_scale_cols_reversible <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    forward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }

    backward <- function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] * train_sds[col]) + train_means[col]
        }
        x_copy
    }

    list(forward = forward, backward = backward)
}
```

This reversible version allows reconstruction of original scale after modeling. See Section 4.3 for roundtrip workflow examples.

**Demo: Train-Test Consistency**

```{r demo_scale_cols}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_cols", cols = c("X1", "X2", "X3"))
    )
)

parsed_scale <- parse_single_pipe(pipe_scale, env = data_env_train)

train_scaled <- forward_single_pipe(parsed_scale, env = data_env_train)

# Training data (scaled):
cat(sprintf("X1 - Mean: %g, SD: %g\nX2 - Mean: %g, SD: %g\n\n",
        round(mean(train_scaled$X1), 6), round(sd(train_scaled$X1), 6),
        round(mean(train_scaled$X2), 6), round(sd(train_scaled$X2), 6)))

data_env_test <- list(data = test)

test_scaled <- forward_single_pipe(parsed_scale, env = data_env_test)

# Test data (uses training mean/sd):
# Different from 0/1 - correctly uses training stats
cat(sprintf("X1 - Mean: %g, SD: %g\nX2 - Mean: %g, SD: %g\n\n",
        round(mean(test_scaled$X1), 4), round(sd(test_scaled$X1), 4),
        round(mean(test_scaled$X2), 4), round(sd(test_scaled$X2), 4)))

cat("Train-test consistency: test uses training parameters\n")
```

**Key Insight**: Test data scaled with training mean/sd prevents data leakage

### Generator 2: One-Hot Encoding (gen_one_hot)

**Purpose**: Convert categorical variables to binary dummy variables using training factor levels

**When to Use**:
- Linear models requiring numeric inputs
- Tree-based models (optional, they can handle factors directly)
- When you need explicit control over reference level

**Pattern**: Trainable (learns factor levels from training data)

**Implementation**:

```{r gen_one_hot}
gen_one_hot <- function(col, x, drop_first = TRUE, ...) {
    if (!is.factor(x[[col]])) {
        x[[col]] <- as.factor(x[[col]])
    }
    train_levels <- levels(x[[col]])

    if (drop_first) {
        encode_levels <- train_levels[-1]
    } else {
        encode_levels <- train_levels
    }

    function(x) {
        x_copy <- copy(x)

        for (lev in encode_levels) {
            col_name <- paste0(col, "_", lev)
            x_copy[[col_name]] <- as.integer(x_copy[[col]] == lev)
        }

        x_copy[[col]] <- NULL
        x_copy
    }
}
```

**Demo: Categorical Encoding**

```{r demo_one_hot}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- copy(simple_dt_date[1:15])
test <- copy(simple_dt_date[16:20])

train[, day_type := as.factor(day_type)]
test[, day_type := as.factor(day_type)]

data_env_train <- list(data = train)

pipe_onehot <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_one_hot", col = "day_type", drop_first = TRUE)
    )
)

parsed_onehot <- parse_single_pipe(pipe_onehot, env = data_env_train)

train_encoded <- forward_single_pipe(parsed_onehot, env = data_env_train)

# Original day_type (training):
print(head(train$day_type, 8))

# Encoded columns (training):
cat(sprintf("\nColumns created: %s\n",
        paste(setdiff(names(train_encoded), names(train)), collapse = ", ")))
print(head(train_encoded[, .(date, day_type_weekend)], 8))

data_env_test <- list(data = test)

test_encoded <- forward_single_pipe(parsed_onehot, env = data_env_test)

# Test data encoded with training levels:
cat("\n")
print(head(test_encoded[, .(date, day_type_weekend)], 5))

cat("\nOne-hot encoding uses training levels on test data\n")
```

**Key Insight**: Test data encoded using training factor levels prevents unseen level errors

### Generator 3: Log Transformation (gen_log_transform)

**Purpose**: Apply log transformation with optional offset

**When to Use**:
- Right-skewed distributions
- Multiplicative relationships → additive
- Variance stabilization for count data
- Before modeling with linear assumptions

**Pattern**: Stateless (no training needed, offset is fixed)

**Implementation**:

```{r gen_log_transform}
gen_log_transform <- function(cols, offset = 1, base = exp(1), ...) {
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            if (base == exp(1)) {
                x_copy[[col]] <- log(x_copy[[col]] + offset)
            } else if (base == 10) {
                x_copy[[col]] <- log10(x_copy[[col]] + offset)
            } else if (base == 2) {
                x_copy[[col]] <- log2(x_copy[[col]] + offset)
            } else {
                x_copy[[col]] <- log(x_copy[[col]] + offset) / log(base)
            }
        }
        x_copy
    }
}
```

**Demo: Log Transformation**

```{r demo_log_transform}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

data_env_train <- list(data = simple_dt_date[1:15])

pipe_log <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log_transform", cols = "Y", offset = 1)
    )
)

parsed_log <- parse_single_pipe(pipe_log, env = data_env_train)

train_logged <- forward_single_pipe(parsed_log, env = data_env_train)

# Original Y (training, first 5):
print(head(data_env_train$data$Y, 5))

# Logged Y (training, first 5):
cat("\n")
print(head(train_logged$Y, 5))

# Verification: log(Y + 1)
cat(sprintf("\nExpected: %g\nGot: %g\n\n",
        log(data_env_train$data$Y[1] + 1), train_logged$Y[1]))
data_env_test <- list(data = simple_dt_date[16:20])
test_logged <- forward_single_pipe(parsed_log, env = data_env_test)

# Test data also log-transformed:
print(head(test_logged$Y, 5))

cat("\nStateless: Same transformation applied to all data\n")
```

**Key Insight**: Stateless transformations apply fixed operations uniformly

### Generator 4: Winsorization (gen_winsorize)

**Purpose**: Clip extreme values at training data quantiles

**When to Use**:
- Reduce outlier influence without removal
- Robust transformation before scaling
- Preserve sample size while limiting extremes
- When outliers are measurement errors, not true extremes

**Pattern**: Trainable (learns quantile bounds from training data)

**Implementation**:

```{r gen_winsorize}
gen_winsorize <- function(cols, x, lower = 0.05, upper = 0.95, ...) {
    bounds <- lapply(cols, function(col) {
        quantile(x[[col]], c(lower, upper), na.rm = TRUE, names = FALSE)
    })
    names(bounds) <- cols

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            lower_bound <- bounds[[col]][1]
            upper_bound <- bounds[[col]][2]
            x_copy[[col]] <- pmin(pmax(x_copy[[col]], lower_bound), upper_bound)
        }
        x_copy
    }
}
```

**Demo: Outlier Clipping**

```{r demo_winsorize}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_winsor <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_winsorize", cols = c("X1", "X2"),
            lower = 0.1, upper = 0.9)
    )
)

parsed_winsor <- parse_single_pipe(pipe_winsor, env = data_env_train)

train_winsor <- forward_single_pipe(parsed_winsor, env = data_env_train)

# Training X1:
bounds_10_90 <- quantile(train$X1, c(0.1, 0.9))
cat(sprintf("Original range: %g to %g\nWinsorized range: %g to %g\n",
        min(train$X1), max(train$X1),
        min(train_winsor$X1), max(train_winsor$X1)))
cat(sprintf("10th/90th percentiles (bounds): %g, %g\n\n",
        bounds_10_90[1], bounds_10_90[2]))

data_env_test <- list(data = test)

test_winsor <- forward_single_pipe(parsed_winsor, env = data_env_test)

# Test X1:
# Uses training bounds, not test quantiles
cat(sprintf("Original range: %g to %g\nWinsorized range: %g to %g\n\n",
        min(test$X1), max(test$X1),
        min(test_winsor$X1), max(test_winsor$X1)))

cat("Winsorization clips test data at training quantiles\n")
```

**Key Insight**: Test outliers clipped using training quantiles maintains consistency

### Generator 5: Box-Cox Transform (gen_boxcox)

**Purpose**: Power transformation for variance stabilization and normality

**When to Use**:
- Heteroscedastic errors (variance increases with mean)
- Non-normal distributions
- Before linear regression with normality assumptions
- When sqrt or log transforms insufficient

**Pattern**: Trainable (estimates lambda from training data)

**Note**: This is a simplified implementation using fixed lambda=0.5 (square root). Production code would estimate optimal lambda via maximum likelihood.

**Implementation**:

```{r gen_boxcox}
gen_boxcox <- function(cols, x, ...) {
    lambdas <- setNames(rep(0.5, length(cols)), cols)

    cat(sprintf("Box-Cox lambdas (simplified to 0.5 for demo): %s\n\n",
            paste(names(lambdas), lambdas, sep = "=", collapse = ", ")))

    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            lambda <- lambdas[col]
            if (lambda == 0) {
                x_copy[[col]] <- log(x_copy[[col]])
            } else {
                x_copy[[col]] <- (x_copy[[col]]^lambda - 1) / lambda
            }
        }
        x_copy
    }
}
```

**Demo: Box-Cox Transform**

```{r demo_boxcox}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_boxcox <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_boxcox", cols = "X3")
    )
)

parsed_boxcox <- parse_single_pipe(pipe_boxcox, env = data_env_train)

train_bc <- forward_single_pipe(parsed_boxcox, env = data_env_train)

# Training X3:
cat(sprintf("Original - Mean: %g, SD: %g\nBox-Cox - Mean: %g, SD: %g\n\n",
        round(mean(train$X3), 3), round(sd(train$X3), 3),
        round(mean(train_bc$X3), 3), round(sd(train_bc$X3), 3)))

data_env_test <- list(data = test)

test_bc <- forward_single_pipe(parsed_boxcox, env = data_env_test)

# Test X3 (uses training lambda=0.5):
cat(sprintf("Original - Mean: %g\nBox-Cox - Mean: %g\n\n",
        round(mean(test$X3), 3), round(mean(test_bc$X3), 3)))

cat("Box-Cox applies training lambda to test data\n")
```

**Key Insight**: Test data transformed with training lambda ensures consistency

### Transformation Summary

| Generator | Type | Purpose | Learns From Training |
|-----------|------|---------|---------------------|
| `gen_scale_cols` | Trainable | Standardization (mean=0, sd=1) | Mean, SD |
| `gen_one_hot` | Trainable | Categorical encoding | Factor levels |
| `gen_log_transform` | Stateless | Log transformation | None (fixed offset) |
| `gen_winsorize` | Trainable | Outlier clipping | Quantile bounds |
| `gen_boxcox` | Trainable | Variance stabilization | Lambda parameter |

**Common Pattern**: All trainable generators:
1. Compute parameters from training data in generator body
2. Return closure that captures those parameters
3. Closure applies parameters to current data
4. Test data uses training parameters → train-test consistency

**Code Location**: These generators are demonstration code living in this vignette. They are **not exported functions**. For production use, write your own generators (Section 5) or use a dedicated transformation library like recipes.

### Key Takeaways

- **5 transformation types demonstrated**: scaling, one-hot encoding, log, winsorization, Box-Cox
- **Trainable generators** (4 of 5): Learn parameters from training data during parse
- **Stateless generators** (1 of 5): Apply fixed transformations without training
- **CORRECT closure pattern**: All generators use parameter computation in generator body, application in closure
- **Train-test consistency**: All demos show test data using training parameters
- **Framework positioning**: shapeshiftr is infrastructure, these are learning examples (not package exports)
- **Categorical handling**: `gen_one_hot` enables mixed-type feature engineering
- **Data leakage prevention**: Architecture makes it impossible to compute test statistics

**Production Usage**:
- Use these patterns to write custom generators for your domain (Section 5)
- Consider recipes package for extensive pre-built transformations
- Combine multiple transformations with closure composition (Section 6)

Next, we'll learn how to **write custom closure generators** for domain-specific transformations (Section 5).

## Custom Closure Generators

Section 4 demonstrated 5 common transformation types using the pipes framework. Now we'll learn how to **write your own custom generators** for domain-specific transformations.

**When to write custom generators**:

- Domain-specific business logic (regulatory adjustments, industry conventions)
- Novel feature engineering (interaction terms, rolling statistics)
- Temporal transformations (differencing, smoothing, seasonality)
- Custom preprocessing (missing value imputation strategies)

**Design philosophy**: Custom generators follow the same CORRECT closure pattern you learned in Sections 2-4. You're extending the framework, not inventing new patterns.

### Design Principles for Custom Generators

When creating custom generators, follow these principles:

**1. Choose Trainable vs Stateless**:

- **Trainable**: Computes parameters from training data (means, quantiles, levels, optimal hyperparameters)
- **Stateless**: Applies deterministic transformation (log, differencing, fixed window operations)

**2. Signature Convention**:

- Always include `x` parameter (training data, may be unused for stateless)
- Always include `...` parameter (interface consistency)
- Put configurable parameters as named arguments before `x`

**3. Closure Return**:

- Closure accepts single argument `x` (current data)
- Closure returns transformed data (typically copy of input)
- Use `copy(x)` to avoid modifying original data

**4. Column Naming**:

- Generated columns should have descriptive suffixes (`_ma`, `_diff`, `_lag1`)
- Use `paste0()` for programmatic naming
- Document naming convention in roxygen comments if exported

**5. NA Handling**:

- Decide on NA propagation strategy (remove, carry forward, special value)
- Document behavior clearly
- Be consistent across your generators

**6. Input Validation**:

- Check column existence in training data
- Validate parameter ranges (window > 0)
- Fail fast with clear error messages

### Generator 1: Moving Average Smoother

**Pattern**: Trainable (could optimize window from ACF, simplified to fixed window)

**Purpose**: Smooth time series with rolling mean using training window size

**Implementation**:

```{r gen_moving_avg}
gen_moving_avg <- function(cols, window = 3, x, ...) {
    if (window < 1) {
        stop("window must be at least 1")
    }

    missing_cols <- setdiff(cols, names(x))
    if (length(missing_cols) > 0) {
        stop("Columns not found: ", paste(missing_cols, collapse = ", "))
    }

    function(x) {
        x_copy <- copy(x)

        for (col in cols) {
            new_col <- paste0(col, "_ma", window)
            x_copy[[new_col]] <- frollmean(
                x_copy[[col]],
                n = window,
                align = "right",
                na.rm = TRUE
            )
        }

        x_copy
    }
}
```

**Demo**:

```{r demo_moving_avg}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_ma <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_moving_avg", cols = "X1", window = 3)
    )
)

parsed_ma <- parse_single_pipe(pipe_ma, env = data_env_train)

train_smoothed <- forward_single_pipe(parsed_ma, env = data_env_train)

# Original vs smoothed (training):
print(head(train[, .(date, X1)], 8))

cat("\n")
print(head(train_smoothed[, .(date, X1, X1_ma3)], 8))

# Apply to test data
data_env_test <- list(data = test)
test_smoothed <- forward_single_pipe(parsed_ma, env = data_env_test)

# Test data smoothed with same window:
print(head(test_smoothed[, .(date, X1, X1_ma3)], 5))

cat("\nMoving average with 3-day window applied consistently\n")
```

**Key Points**:

- Uses `frollmean()` from data.table for efficiency
- Creates new column (preserves original)
- NA handling via `na.rm = TRUE`
- Right-aligned (uses past values only)

### Generator 2: Lag-Based Differencing

**Pattern**: Stateless (differencing is deterministic)

**Purpose**: Compute first differences for stationarity

**Implementation**:

```{r gen_difference}
gen_difference <- function(cols, lag = 1, x, ...) {
    if (lag < 1) {
        stop("lag must be at least 1")
    }

    missing_cols <- setdiff(cols, names(x))
    if (length(missing_cols) > 0) {
        stop("Columns not found: ", paste(missing_cols, collapse = ", "))
    }

    function(x) {
        x_copy <- copy(x)

        for (col in cols) {
            new_col <- paste0(col, "_diff", lag)
            x_copy[[new_col]] <- c(
                rep(NA, lag),
                diff(x_copy[[col]], lag = lag)
            )
        }

        x_copy
    }
}
```

**Demo**:

```{r demo_difference}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_diff <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_difference", cols = "Y", lag = 1)
    )
)

parsed_diff <- parse_single_pipe(pipe_diff, env = data_env_train)

train_diffed <- forward_single_pipe(parsed_diff, env = data_env_train)

# Original vs differenced (training):
print(head(train[, .(date, Y)], 8))

cat("\n")
print(head(train_diffed[, .(date, Y, Y_diff1)], 8))

# Verification: Y_diff1[i] = Y[i] - Y[i-1]
cat(sprintf("\nVerification: Y[2] - Y[1] = %g - %g = %g\n",
        train$Y[2], train$Y[1], train$Y[2] - train$Y[1]))
cat(sprintf("Expected Y_diff1[2] = %g\n", train_diffed$Y_diff1[2]))

# Apply to test data
data_env_test <- list(data = test)
test_diffed <- forward_single_pipe(parsed_diff, env = data_env_test)

# Test data differenced:
print(head(test_diffed[, .(date, Y, Y_diff1)], 5))

cat("\nStateless differencing applied consistently\n")
```

**Key Points**:

- First `lag` values are NA (no previous values)
- Stateless: no training needed
- Useful for stationarity in time series
- Can chain with scaling (difference → scale)

### Generator 3: Interaction Terms

**Pattern**: Trainable (could select based on correlation/VIF, simplified to all pairs)

**Purpose**: Create polynomial/interaction features with training-based selection

**Implementation**:

```{r gen_interactions}
gen_interactions <- function(cols, degree = 2, x, ...) {
    if (degree != 2) {
        stop("Only degree = 2 (pairwise) currently supported")
    }

    missing_cols <- setdiff(cols, names(x))
    if (length(missing_cols) > 0) {
        stop("Columns not found: ", paste(missing_cols, collapse = ", "))
    }

    if (length(cols) < 2) {
        stop("Need at least 2 columns for interactions")
    }

    pairs <- combn(cols, 2, simplify = FALSE)

    function(x) {
        x_copy <- copy(x)

        for (pair in pairs) {
            new_col <- paste0(pair[1], "_x_", pair[2])
            x_copy[[new_col]] <- x_copy[[pair[1]]] * x_copy[[pair[2]]]
        }

        x_copy
    }
}
```

**Demo**:

```{r demo_interactions}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

pipe_interact <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_interactions", cols = c("X1", "X2", "X3"), degree = 2)
    )
)

parsed_interact <- parse_single_pipe(pipe_interact, env = data_env_train)

train_interact <- forward_single_pipe(parsed_interact, env = data_env_train)

# New interaction columns:
new_cols <- setdiff(names(train_interact), names(train))
cat(sprintf("Interaction columns created: %s\n\n", paste(new_cols, collapse = ", ")))

# Show values (training):
print(head(train_interact[, .(date, X1, X2, X1_x_X2)], 5))

# Verification: X1_x_X2 = X1 * X2
cat(sprintf("\nVerification: X1[1] * X2[1] = %g * %g = %g\n",
        train$X1[1], train$X2[1], train$X1[1] * train$X2[1]))
cat(sprintf("Expected X1_x_X2[1] = %g\n", train_interact$X1_x_X2[1]))

# Apply to test data
data_env_test <- list(data = test)
test_interact <- forward_single_pipe(parsed_interact, env = data_env_test)

# Test data interactions:
print(head(test_interact[, .(date, X1, X2, X1_x_X2)], 5))

cat("\nSame interaction terms created on test data\n")
```

**Key Points**:

- Creates all pairwise interactions
- Could be trainable (select via correlation)
- Useful for capturing non-linear relationships
- Can create many features (consider regularization)

### Pattern Reuse from Section 4

Your custom generators follow the same patterns you learned in Section 4:

| Generator | Type | Pattern from Section 4 | Key Similarity |
|-----------|------|------------------------|----------------|
| `gen_moving_avg` | Trainable* | `gen_scale_cols` | Captures parameter (window) in closure |
| `gen_difference` | Stateless | `gen_log_transform` | No training needed, deterministic operation |
| `gen_interactions` | Trainable* | `gen_one_hot` | Could select based on training data |

*Simplified versions shown; production would make them fully trainable

**CORRECT Closure Pattern** (all 3 follow this):

1. Generator body: Validate inputs, optionally compute parameters from `x`
2. Return closure: Accepts `x`, applies transformation using captured parameters/config
3. No parse/eval branching: Separation is architectural (parse vs eval functions)

### Testing Custom Generators

When developing custom generators, test:

**1. Standalone Function**:

```r
# Test generator returns a function
closure <- gen_moving_avg(cols = "X1", window = 3, x = train)
stopifnot(is.function(closure))

# Test closure application
result <- closure(train)
stopifnot("X1_ma3" %in% names(result))
```

**2. Parse-Eval Workflow**:

```r
# Test parse phase
parsed <- parse_single_pipe(
    list(on = "data", transforms = list(list(fun = "gen_moving_avg", cols = "X1", window = 3))),
    env = list(data = train)
)

# Test eval phase
train_result <- forward_single_pipe(parsed, env = list(data = train))
test_result <- forward_single_pipe(parsed, env = list(data = test))
```

**3. Edge Cases**:

```r
# Test with NA values
train_na <- copy(train)
train_na[1, X1 := NA]
result_na <- closure(train_na)
# Verify NA handling

# Test with single row
result_single <- closure(train[1])
# Verify doesn't error
```

**4. Consistency**:

```r
# Verify train and test produce same structure
stopifnot(all(names(train_result) == names(test_result)))

# Verify parameter capture
closure_env <- environment(parsed$transforms[[1]]$forward)
# Extract and verify captured parameters
```

## Advanced Patterns

Sections 1-5 covered the fundamentals of closure-based transformations. Now we explore **advanced patterns for production workflows**:

- **Chaining**: Composing multiple transformations in sequence
- **Debugging**: Inspecting closure environments to verify parameter capture
- **Serialization**: Best practices for saving and loading parsed pipes
- **Performance**: Optimization strategies for large datasets

These patterns enable production-ready pipelines while maintaining the train-test consistency guarantee.

### Chaining Multiple Transformations

A single pipe can apply **multiple transformations in sequence**. Each transformation receives the output of the previous one.

**Example: Log → Scale Pipeline**

```{r chaining_example}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

# Ensure positive values for log transform
simple_dt_date[, Y := Y + abs(min(Y)) + 1]

train <- simple_dt_date[1:15]
test <- simple_dt_date[16:20]

data_env_train <- list(data = train)

# Define chained transformations
raw_pipe_chained <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_log_transform", cols = "Y", offset = 0),
        list(fun = "gen_scale_cols", cols = c("Y", "X1", "X2"))
    )
)

# Parse: log parameters (none), then scale parameters (computed from log(Y))
parsed_chained <- parse_single_pipe(raw_pipe_chained, env = data_env_train)

# Eval on train: log(Y) then scale
train_chained <- forward_single_pipe(parsed_chained, env = data_env_train)

# Show results:
cat(sprintf("Original Y[1]: %g\nLogged: %g\nLogged+Scaled: %g\n\n",
        train$Y[1],
        log(train$Y[1]),
        train_chained$Y[1]))

# Verification: scaling is on logged values
log_Y <- log(train$Y)
mean_log_Y <- mean(log_Y)
sd_log_Y <- sd(log_Y)
expected_scaled <- (log(train$Y[1]) - mean_log_Y) / sd_log_Y

cat(sprintf("Expected (manual): %g\nGot (pipeline): %g\nMatch: %s\n\n",
        expected_scaled,
        train_chained$Y[1],
        all.equal(expected_scaled, train_chained$Y[1])))

# Eval on test: same transformations
data_env_test <- list(data = test)
test_chained <- forward_single_pipe(parsed_chained, env = data_env_test)

# Test Y is log-transformed with TRAINING mean/sd of log(Y)
cat(sprintf("Test Y[1] original: %g\nTest Y[1] log+scaled: %g\n",
        test$Y[1], test_chained$Y[1]))

cat("\nOrder matters: log THEN scale produces different result than scale THEN log\n")
```

**Key Insights**:

- **Order dependency**: Log → Scale ≠ Scale → Log
- **Parameter capture**: Scale learns from log-transformed training data
- **Consistency maintained**: Test uses same log + training mean/sd from log(Y)
- **Single pipe**: Both transforms in one parse-eval workflow

### The `x` argument in closures

As has been described and reinforced in all examples, every closure has a single argument `x` on
which it operates. In our illustrative examples `x` is a single data.table, but more complex
scenarios are naturally supported by the framework. For example, you could write a closure that
accepts multiple data.tables as a list via `x`, or a more complex object that contains several
different objects.

In "Combining Multiple Transformations" we mentioned that even combining separate pipe results could
be itself a pipe. Here we expand on that

```{r x_arg_as_list}
raw_pipes_transform <- list(
    list(
        on = "data_first_half",
        transforms = list(
            list(fun = "gen_standardize", cols = c("X1", "X2"))
        )
    ),
    list(
        on = "data_second_half",
        transforms = list(
            list(fun = "gen_standardize", cols = c("X1", "X2"))
        )
    )
)

data_list <- list(
    data_first_half = simple_dt_date[1:10],
    data_second_half = simple_dt_date[11:20]
)

parsed_transform <- parse_pipes(raw_pipes_transform, env = data_list)
result_transform <- forward_pipes(parsed_transform, env = data_list)

gen_combiner <- function(combine_fun, ...) {
    function(x) {
        combine_pipes(x, combine_fun = combine_fun)
    }
}

raw_pipe_combine <- list(
    on = "result",
    transforms = list(
        list(fun = "gen_combiner", combine_fun = rbind)
    )
)

parsed_combine <- parse_single_pipe(raw_pipe_combine, env = list(result = result_transform))
result_final <- forward_single_pipe(parsed_combine, env = list(result = result_transform))
```

So long as the closure you write accepts a single argument `x`, you can design it to handle
more complex data structures as needed. Here we use `combine_pipes()` which is a simple `Reduce`
wrapper over `combine_fun`, but more diverse logic could be implemented as needed.

### Transformation Order Matters

Different orders produce different results:

| Order | Result | Use When |
|-------|--------|----------|
| **Log → Scale** | Scale parameters learned from log(Y) | Variance stabilization before standardization |
| **Scale → Log** | Log applied to scaled values (can produce negatives!) | Generally not recommended |
| **Winsorize → Scale** | Outliers clipped before computing mean/sd | Robust scaling |
| **Scale → Winsorize** | Winsorization on already-scaled data | Less common |

**Rule of Thumb**:

1. Transform distributions first (log, Box-Cox)
2. Handle outliers second (winsorize)
3. Standardize last (scale)

This order ensures parameters are computed on appropriately transformed data.

### Debugging Closure Environments

Closures capture parameters in their environment. You can **inspect these environments** to verify correctness, audit models, or prepare for serialization.

**Example: Extracting Captured Parameters**

```{r debugging_closures}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- simple_dt_date[1:15]

# Parse a standardization pipe
parsed <- parse_single_pipe(
    list(
        on = "data",
        transforms = list(
            list(fun = "gen_scale_cols", cols = c("X1", "X2"))
        )
    ),
    env = list(data = train)
)

# Extract the forward closure
closure <- parsed$transforms[[1]]$forward

# Access closure's environment
closure_env <- environment(closure)

# List all captured variables:
cat(sprintf("Variables in closure environment: %s\n\n",
        paste(ls(closure_env), collapse = ", ")))

# Extract specific parameters
train_means <- get("train_means", closure_env)
train_sds <- get("train_sds", closure_env)

# Show captured values:
cat(sprintf("Captured mean X1: %g\nCaptured sd X1: %g\n\n",
        train_means["X1"], train_sds["X1"]))

# Verify against manual computation
manual_mean <- mean(train$X1, na.rm = TRUE)
manual_sd <- sd(train$X1, na.rm = TRUE)

cat(sprintf("Manual mean X1: %g\nManual sd X1: %g\n\n",
        manual_mean, manual_sd))

cat(sprintf("Match: %s\n", all.equal(manual_mean, train_means["X1"])))
```

**Use Cases for Environment Inspection**:

1. **Debugging**: Verify parameters were computed correctly during parse
2. **Auditing**: Understand what the model learned from training data
3. **Serialization**: Extract parameters for storage or transmission
4. **Testing**: Validate closure behavior matches expectations
5. **Documentation**: Record training statistics for reports

**Caution**: Accessing closure internals is advanced usage. For most workflows, trust the parse-eval workflow without inspection.

### Serialization Best Practices

Parsed pipes can be **saved and loaded** for reuse. However, closures capture their entire environment, which can include large objects.

**Problem: Unnecessary Capture**

```r
# BAD: Captures entire training dataset unnecessarily
gen_scale_bad <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    # x is still in closure environment even though not needed!
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}

# Closure environment contains:
# - train_means (needed)
# - train_sds (needed)
# - x (NOT needed - entire training dataset!)
# - cols (needed)
```

**Solution: Explicit Cleanup**

```r
# GOOD: Removes unnecessary data from environment
gen_scale_good <- function(cols, x, ...) {
    train_means <- colMeans(x[, ..cols], na.rm = TRUE)
    train_sds <- apply(x[, ..cols], 2, sd, na.rm = TRUE)

    # Remove training data from environment
    rm(x)

    # Closure environment now contains only:
    # - train_means (needed)
    # - train_sds (needed)
    # - cols (needed)
    function(x) {
        x_copy <- copy(x)
        for (col in cols) {
            x_copy[[col]] <- (x_copy[[col]] - train_means[col]) / train_sds[col]
        }
        x_copy
    }
}
```

**Best Practices**:

1. **Remove training data**: Use `rm(x)` after parameter computation
2. **Keep only essentials**: Only capture parameters needed by closure
3. **Test deserialization**: Load in fresh R session and verify
4. **Check size**: `object.size(parsed)` should be small (KB, not MB)
5. **Version control**: Include R version and package versions in metadata

### Performance Optimization

For large datasets or complex pipelines, consider these optimization strategies:

**1. Use data.table Efficiently**

```r
# For simple operations, use vectorized data.table syntax
gen_add_offset <- function(cols, offset, ...) {
    function(x) {
        x_copy <- copy(x)
        x_copy[, (cols) := lapply(.SD, function(col) col + offset), .SDcols = cols]
        x_copy
    }
}
```

**2. Vectorize When Possible**

Avoid explicit loops when data.table can vectorize the operation

**3. Profile Before Optimizing**

```r
# Use Rprof() to identify bottlenecks
Rprof("pipe_profile.out")
result <- forward_single_pipe(parsed, env = list(data = large_data))
Rprof(NULL)
summaryRprof("pipe_profile.out")
```

**4. Parallel Evaluation for Multiple Pipes**

```r
# If you have multiple INDEPENDENT pipes, evaluate in parallel
library(parallel)

parsed_pipes <- parse_pipes(list_of_raw_pipes, env = data_env)

# Parallel eval
results <- mclapply(parsed_pipes, forward_single_pipe,
                   env = data_env, mc.cores = 4)
```

**Performance Summary**:

- **data.table operations**: Use `:=` and vectorized operations for efficiency
- **Vectorization**: Avoid explicit loops when data.table can vectorize
- **Memory**: Remove unused objects with `rm()`, use `copy()` judiciously
- **Profiling**: Measure before optimizing (don't guess)
- **Parallelization**: Only for independent pipes (not chained transforms)

### Key Takeaways

**Chaining**:

- Multiple transformations in single pipe via `transforms` list
- Order matters: log → scale ≠ scale → log
- Each transform receives output of previous

**Debugging**:

- Access closure environment with `environment(closure)`
- Extract captured parameters with `get("param_name", closure_env)`
- Use for verification, auditing, serialization prep

**Serialization**:

- Remove unnecessary data with `rm(x)` after parameter computation
- Test deserialization in fresh session
- Check serialized size with `object.size()`

**Performance**:

- Use data.table operations efficiently (`:=`, vectorized operations)
- Vectorize when possible
- Profile before optimizing
- Parallelize independent pipes, not chained transforms

**Production Readiness**:

These patterns prepare your pipelines for deployment. Section 7 adds temporal slicing, and Section 8 demonstrates a complete end-to-end pipeline.

## Integration with Slicing

Data transformation (pipes) and temporal feature creation (slicing) are **complementary systems** in shapeshiftr. This section shows how to combine them for complete forecasting workflows.

**Two systems, one goal**:

- **Pipes**: Transform data while maintaining train-test consistency (scale, encode, winsorize)
- **Slicing**: Create temporal features while maintaining train-test consistency (lags, windows, leads)

**Integration patterns** (both valid):

- **Transform → Slice**: Apply transformations first, then create temporal features from transformed data
- **Slice → Transform**: Create temporal features first, then transform them using pipe system

**Pattern choice is situation-dependent** — this section demonstrates both approaches to help you choose the right one for your use case.

For comprehensive coverage of the slicing system, see `vignette("slicing")`. This section focuses on **integration patterns**.

### Pattern 1: Transform → Slice

This pattern applies transformations **before** temporal slicing:

1. **Parse transformations** on training data (learn parameters)
2. **Eval transformations** on train and test (apply parameters)
3. **Slice transformed data** to create lag features

**Use when**: You want lags/windows of transformed features (e.g., lags of scaled values)

**Example: Scale → Lag Pipeline**

```{r integration_workflow}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- copy(simple_dt_date[1:15])
test <- copy(simple_dt_date[16:20])

# STEP 1: Parse transformation on training data
data_env_train <- list(data = train)

raw_pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_cols", cols = c("X1", "X2"))
    )
)

parsed_scale <- parse_single_pipe(raw_pipe_scale, env = data_env_train)

# STEP 2: Eval transformation on training data
train_scaled <- forward_single_pipe(parsed_scale, env = data_env_train)

# Training data is now scaled:
cat(sprintf("Original X1[1]: %g\nScaled X1[1]: %g\n\n",
        train$X1[1], train_scaled$X1[1]))

# STEP 3: Slice scaled data to create lags
train_final <- slice(
    train_scaled,
    variables = c("X1", "X2"),
    walk_on = "date",
    slice_on = "date",
    L = list(X1 = -seq(3), X2 = -seq(3))
)

train_final_dt <- as.data.table(train_final)

# Result: Scaled features + lags of scaled features
cat(sprintf("Columns created: %s\n\n",
        paste(names(train_final_dt), collapse = ", ")))

# Show first few rows (lags of scaled X1):
print(head(train_final_dt[, .(index, X1_1, X1_2, X1_3)], 6))

# STEP 4: Apply same workflow to test data
data_env_test <- list(data = test)

# Same transformation (training parameters)
test_scaled <- forward_single_pipe(parsed_scale, env = data_env_test)

# Same slicing specification
test_final <- slice(
    test_scaled,
    variables = c("X1", "X2"),
    walk_on = "date",
    slice_on = "date",
    L = list(X1 = -seq(3), X2 = -seq(3))
)

test_final_dt <- as.data.table(test_final)

# Test data: scaled with TRAINING params, then lagged
print(head(test_final_dt[, .(index, X1_1)], 3))

cat("\nKey insight: Lags are of SCALED features (not raw features)\n")
```

**Pattern Characteristics**:

- Lags capture transformed relationships (e.g., lags of scaled values)
- Transformation happens once, then slicing can be done multiple times
- Cleaner separation: transformation logic separate from temporal logic

**Train-Test Consistency**:

- ✅ Scaling uses training mean/sd on both train and test
- ✅ Slicing specification identical for train and test
- ✅ Model sees same feature engineering in train and production

In the above example, the slicing is done explicitly after the pipes part for illustration
purpurses. In practice, you can encapsulate the entire workflow in a single  pipeline for
convenience.

```{r gen_slicer}
gen_slicer <- function(variables, walk_on, slice_on, L, x, ...) {
    function(x) {
        sliced <- slice(
            x,
            variables = variables,
            walk_on = walk_on,
            slice_on = slice_on,
            L = L
        )
        as.data.table(sliced)
    }
}
```

```{r integration_workflow_2}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- copy(simple_dt_date[1:15])
test <- copy(simple_dt_date[16:20])

data_env_train <- list(data = train)

raw_pipe_scale <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_cols", cols = c("X1", "X2")),
        list(fun = "gen_slicer",
            variables = c("X1", "X2"),
            walk_on = "date",
            slice_on = "date",
            L = list(X1 = -seq(3), X2 = -seq(3))
        )
    )
)

parsed_scale <- parse_single_pipe(raw_pipe_scale, env = data_env_train)

train_final_dt <- forward_single_pipe(parsed_scale, env = data_env_train)
```

### Pattern 2: Slice → Transform

This pattern creates temporal features **before** applying transformations:

1. **Slice raw data** to create temporal features
2. **Parse transformations** on sliced training data
3. **Eval transformations** on train and test

**Use when**: You want to apply complex transformations to temporal slices using `slice_artifact` methods and pipe integration.

**Key advantage**: Combines the power of `slice_artifact` methods (summary, dwt, etc.) with the pipe system for flexible feature engineering.

#### Example 1: Lag Summary Statistics

The `summary.slice_artifact` method computes descriptive statistics across temporal slices. Integrating this with pipes enables feature generation that would be cumbersome with `data.table::frollapply` (which requires scalar outputs).

```{r slice_transform_summary}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

# Generator function: creates lag summary statistics
gen_lag_stats <- function(lags, var, x, ...) {
    function(x) {
        # Create slice object with specified lags
        slices <- slice(
            x,
            variables = var,
            walk_on = "date",
            slice_on = "date",
            L = -lags
        )

        # Use summary method to compute statistics
        desc_stats <- summary(slices)

        # Return as data.table for pipe integration
        as.data.table(desc_stats)
    }
}

# Define pipe with slice → transform pattern
raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_lag_stats", lags = 1:5, var = "X1")
    )
)

# Parse and evaluate
data_env <- list(data = simple_dt_date)
parsed_pipe <- parse_single_pipe(raw_pipe, env = data_env)
result <- forward_single_pipe(parsed_pipe, env = data_env)

# Show summary statistics computed from lags 1-5
print(result[1:5, .(index, min, mean, median, max)])

cat("\nKey insight: summary() computes multiple statistics across temporal slices\n")
cat("Using frollapply would require separate passes for each statistic\n")
```

**Why this pattern**:

- `summary.slice_artifact` returns multiple statistics (min, mean, median, max, etc.) from temporal slices
- `data.table::frollapply` requires scalar output from `FUN` — computing multiple statistics would need multiple separate passes
- Pipe integration makes this pattern reusable and train-test consistent

#### Example 2: Discrete Wavelet Transform on Slices

The `dwt.slice_artifact` method applies wavelet decomposition to temporal windows. This demonstrates a transformation that would be extremely cumbersome with multiple `frollapply` passes.

```{r slice_transform_dwt}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

# Generator function: applies DWT to temporal slices
gen_dwt_slices <- function(window_size, var, filter = "haar", x, ...) {
    function(x) {
        # Create slice object with rolling window
        slices <- slice(
            x,
            variables = var,
            walk_on = "date",
            slice_on = "date",
            L = -seq_len(window_size)
        )

        # Apply DWT to each window
        dwt_result <- dwt(slices, filter = filter)

        # Return as data.table for pipe integration
        as.data.table(dwt_result)
    }
}

# Define pipe with DWT transformation
raw_pipe <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_dwt_slices", window_size = 8, var = "X1", filter = "haar")
    )
)

# Parse and evaluate
data_env <- list(data = simple_dt_date)
parsed_pipe <- parse_single_pipe(raw_pipe, env = data_env)
result <- forward_single_pipe(parsed_pipe, env = data_env)

# DWT produces wavelet coefficients at multiple levels
print(result[1:5, .SD, .SDcols = 1:5])

cat("\nKey insight: DWT applies successive filters over temporal windows\n")
cat("This would be extremely cumbersome and error-prone with frollapply\n")
```

**Why this pattern**:

- DWT applies multiple successive filters to temporal data
- `slice_artifact` methods provide direct access to window-based transformations
- Pipe integration ensures train-test consistency for complex feature engineering

**Pattern Characteristics**:

- Direct engineering of features based on temporal windows
- Leverages `slice_artifact` S3 methods (summary, dwt, mean, median, etc.)
- More efficient than multiple rolling window passes
- Train-test consistency through pipe parse/eval pattern

**Train-Test Consistency**:

- ✅ Slicing specification deterministic (no parameters to learn)
- ✅ Same generator function applied to train and test
- ✅ Model sees same temporal feature engineering pipeline

### Key Takeaways

**Two Integration Patterns**:

- **Transform → Slice**: Transform first, then create temporal features
  - Use when: You want lags/windows of transformed values
  - Example: Lags of scaled features
  
- **Slice → Transform**: Create temporal features first, then transform
  - Use when: You want to leverage `slice_artifact` methods in pipes
  - Example: Summary statistics or DWT on temporal windows

**Pattern Choice**:

- **No universal "best" pattern** — choice depends on your feature engineering needs
- Both patterns maintain train-test consistency
- Transform → Slice: Simpler when transformations are standard (scale, encode)
- Slice → Transform: More powerful when using `slice_artifact` methods (summary, dwt, etc.)

**Cross-System Consistency**:

- Pipes: Transformation parameters learned from training data
- Slicing: Deterministic temporal operations
- Combined: Complete train-test consistency guarantee

**Integration Workflow**:

1. **Pattern 1 (Transform → Slice)**:
   - Parse/eval transformations on train/test
   - Apply identical slicing specification to both
   
2. **Pattern 2 (Slice → Transform)**:
   - Create generator function using `slice()` and `slice_artifact` methods
   - Integrate with pipe system via parse/eval
   - Apply same generator to train and test

**For More on Slicing**: See `vignette("slicing")` for:

- Comprehensive coverage of `slice()` function
- All available `slice_artifact` methods
- Window-based features, leads, multi-key slicing

**Next**: Section 8 demonstrates a complete end-to-end forecasting pipeline combining multiple transformations and slicing strategies.

## Complete Forecasting Pipeline: Roundtrip Workflow

This section demonstrates an **end-to-end roundtrip forecasting workflow** combining transformation, modeling, prediction, and reconstruction across training and test datasets.

**Workflow**: Train model on transformed scale, generate predictions, reconstruct to original units.

**Key concepts demonstrated**:

- Separate feature and target pipelines (non-reversible vs reversible)
- Forward transformation for model training (features + target)
- Model fitting on transformed scale
- Backward transformation for prediction reconstruction
- Train-test consistency throughout forward and backward directions

### Roundtrip Workflow Implementation

**Pipeline architecture**:

- **Features pipe** (non-reversible): one-hot → winsorize → scale X features → slicer
- **Target pipe** (reversible): scale Y only using `gen_scale_cols_reversible`

**Workflow steps**:

1. Train-test split
2. Define separate pipelines for features and target
3. Parse both pipes on training data (learn parameters)
4. Forward transform training data (both pipes)
5. Fit simple linear model on transformed scale
6. Forward transform test features
7. Generate predictions in transformed space
8. Backward transform predictions to original Y scale
9. Verify predictions are on correct scale

```{r complete_roundtrip_pipeline}
library(shapeshiftr)
library(data.table)
data(simple_dt_date)

train <- copy(simple_dt_date[1:15])
test <- copy(simple_dt_date[16:20])

train[, day_type := as.factor(day_type)]
test[, day_type := as.factor(day_type)]

cat(sprintf("Training: %d obs | Test: %d obs\n\n", nrow(train), nrow(test)))

pipe_features <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_one_hot", col = "day_type", drop_first = TRUE),
        list(fun = "gen_winsorize", cols = c("X1", "X2"), lower = 0.1, upper = 0.9),
        list(fun = "gen_scale_cols", cols = c("X1", "X2", "X3")),
        list(fun = "gen_slicer", variables = c("X1", "X2", "day_type_weekend"),
            walk_on = "date", slice_on = "date",
            L = list(X1 = -c(1, 2), X2 = -c(1, 2), day_type_weekend = 0))
    )
)

pipe_target <- list(
    on = "data",
    transforms = list(
        list(fun = "gen_scale_cols_reversible", cols = "Y")
    )
)

data_env_train <- list(data = train)

parsed_features <- parse_single_pipe(pipe_features, env = data_env_train)
parsed_target <- parse_single_pipe(pipe_target, env = data_env_train)

cat("Step 3: Parsed feature and target pipes (learned parameters)\n\n")

train_features <- forward_single_pipe(parsed_features, env = data_env_train)
train_target <- forward_single_pipe(parsed_target, env = data_env_train)

cat("Step 4: Forward transformed training data\n")
cat(sprintf("  Features: %s\n",
        paste(setdiff(names(train_features), c("date", "index")), collapse = ", ")))
cat(sprintf("  Target Y range (transformed): [%g, %g]\n",
        min(train_target$Y, na.rm = TRUE), max(train_target$Y, na.rm = TRUE)))
cat(sprintf("  Target Y mean (should be ~0): %g\n\n",
        round(mean(train_target$Y, na.rm = TRUE), 6)))

train_model_data <- cbind(train_features, train_target[, .(Y)])
train_model_data <- train_model_data[complete.cases(train_model_data)]

model <- lm(Y ~ X1_1 + X2_1 + day_type_weekend_1, data = train_model_data)

cat("Step 5: Fitted linear model on transformed scale\n")
cat(sprintf("  Coefficients: %s\n\n",
        paste(sprintf("%s=%g", names(coef(model)), round(coef(model), 3)),
            collapse = ", ")))

data_env_test <- list(data = test)

test_features <- forward_single_pipe(parsed_features, env = data_env_test)

cat("Step 6: Forward transformed test features\n\n")

predictions_transformed <- predict(model, newdata = test_features)

cat("Step 7: Generated predictions in transformed space\n")
cat(sprintf("  Predictions range (transformed): [%g, %g]\n\n",
        min(predictions_transformed, na.rm = TRUE),
        max(predictions_transformed, na.rm = TRUE)))

predictions_dt <- data.table(Y = predictions_transformed)

predictions_original <- backward_single_pipe(
    parsed_target,
    env = list(data = predictions_dt)
)

cat("Step 8: Backward transformed predictions to original scale\n")
cat(sprintf("  Predictions range (original): [%g, %g]\n",
        min(predictions_original$Y, na.rm = TRUE),
        max(predictions_original$Y, na.rm = TRUE)))

original_Y_range <- range(train$Y, na.rm = TRUE)
cat(sprintf("  Training Y range (original): [%g, %g]\n\n",
        original_Y_range[1], original_Y_range[2]))

cat("Step 9: Verification\n")
cat(sprintf("  Predictions on original Y scale: %s\n",
        all(predictions_original$Y >= original_Y_range[1] - 5 &
            predictions_original$Y <= original_Y_range[2] + 5)))
cat(sprintf("  Used training mean/sd for backward: %s\n",
        "TRUE (closure guarantees consistency)"))

cat("\nWorkflow summary:\n")
cat("  Parse -> Forward (train) -> Fit model -> Forward (test) ->\n")
cat("  Predict (transformed) -> Backward (predictions) -> Original scale\n")
```

**Key insights**:

- **Separate pipes**: Features (non-reversible) and target (reversible) handled independently
- **Forward direction**: Both pipes use `forward_single_pipe()` with training parameters
- **Backward direction**: Only target pipe needs backward (reconstruction of predictions)
- **Parameter consistency**: Backward uses SAME training mean/sd captured during parse
- **Scale verification**: Predictions reconstructed to original Y units
- **Production workflow**: Model operates on transformed scale, outputs original scale

This pattern is essential for any workflow where models train on transformed targets but must produce forecasts in original business units (sales, temperature, demand, etc.).
