---
title: "The Slicing System: Temporal Feature Engineering for Forecasting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Slicing System: Temporal Feature Engineering for Forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

```{r libraries, message = FALSE}
library(shapeshiftr)
library(data.table)
```

## Introduction

Time series forecasting requires transforming sequential data into a format suitable for
supervised learning algorithms. This transformation involves creating features from past
observations (lags) and aligning them with prediction targets. Doing this manually is
tedious and error-prone.

Consider a simple forecasting scenario where you want to predict tomorrow's value using
the past three days. Manually creating lag features requires careful indexing:

```{r manual-lag-example, echo = FALSE}
manual_example <- data.table(
    date = as.Date(c("2025-01-03", "2025-01-04", "2025-01-05", "2025-01-06")),
    value = c(15, 11, 18, 14),
    value_lag1 = c(12, 15, 11, 18),
    value_lag2 = c(10, 12, 15, 11),
    value_lag3 = c(NA, 10, 12, 15)
)
print(manual_example)
```

This manual approach has several problems:

- **Error-prone**: Easy to misalign rows when creating multiple lags
- **Tedious**: Each variable requires separate lag column creation
- **Inflexible**: Changing lag structure requires rewriting code
- **No reusability**: Cannot easily apply same structure to new data

The **shapeshiftr** package automates this entire process through its **slicing system**.
A single `slice()` call creates all lag features with proper temporal alignment, handles
missing values correctly, and produces model-ready datasets.

### What You Will Learn

This vignette teaches you the complete slicing system for temporal feature engineering:

1. **Introduction** (this section) - Why temporal slicing matters
2. **Simple Slicing** - Creating lag features from the same temporal index
3. **Keyed Slicing** - Integrating data with different temporal resolutions
4. **Building Model-Ready Datasets** - Complete workflow with merge and conversion
5. **Handling Missing Values** - Strategies for incomplete data (future sprint)
6. **Multiple Targets** - Forecasting multiple variables simultaneously (future sprint)
7. **Advanced Patterns** - Complex temporal relationships (future sprint)
8. **Best Practices** - Production guidelines and performance tips (future sprint)

This sprint covers sections 1-4, providing the foundation for all temporal feature
engineering workflows in shapeshiftr.

### Vignette Structure

Sections 2-3 introduce the two core slicing modes (**simple** and **keyed**), while
Section 4 demonstrates the essential workflow for combining these slicing results into
final model-ready datasets. By the end of Section 4, you will be able to build complete
forecasting feature matrices ready for any modeling framework.


## Simple Slicing

**Simple slicing** occurs when you walk and slice on the same temporal index. This is
the most common scenario in forecasting: creating lag features where past values of
variables become predictors for future values.

The defining characteristic is: **`walk_on == slice_on`**

This creates autoregressive features where each row in the result represents a prediction
task, with columns containing lagged values of your variables.

### Concept: Walking and Slicing on the Same Index

When you call `slice()` with the same column for `walk_on` and `slice_on`, shapeshiftr:

1. **Walks** through each time point in your data (controlled by `walk_on`)
2. **Slices** backward (or forward) in time from that point (controlled by `slice_on`)
3. **Extracts** values at the specified lag positions (controlled by `L`)

The result is a set of lag features properly aligned with each time point.

### Example 1: Single Lag (AR(1) Model)

Let's start with the simplest case: creating a single lag feature for forecasting.
This represents an AR(1) autoregressive model where tomorrow's value is predicted
using today's value.

```{r simple-lag-1}
data(simple_dt_date)

lag_1 <- slice(
    simple_dt_date,
    variables = "Y",
    walk_on = "date",
    slice_on = "date",
    L = c(-1)
)

ar1_data <- as.data.table(lag_1)
head(ar1_data)
```

**Key observations**:

- The `walk_on` and `slice_on` are both `"date"` - this is simple slicing
- `L = c(-1)` creates a 1-day lag of variable `Y` (negative = past values)
- The result has columns: `index` (the date), `Y_1` (lag 1 of Y), and `Y` (the target)
- First row has `NA` for `Y_1` (no previous value exists)
- Each subsequent row contains yesterday's `Y` value as `Y_1`

This dataset is immediately ready for fitting an AR(1) model:
`Y ~ Y_1`

### Example 2: Multiple Lags (AR(3) Model)

For richer temporal patterns, we often need multiple lags. An AR(3) model uses
the past 3 values to predict the next value.

```{r simple-lag-3}
lag_3 <- slice(
    simple_dt_date,
    variables = "Y",
    walk_on = "date",
    slice_on = "date",
    L = c(-1, -2, -3)
)

ar3_data <- as.data.table(lag_3)
head(ar3_data)
```

**Key observations**:

- `L = c(-1, -2, -3)` creates lags 1, 2, and 3 simultaneously
- Result has: `index`, `Y_1`, `Y_2`, `Y_3`, `Y`
- First 3 rows have progressively more `NA` values
- Row 4 onward has complete lag features

This pattern extends naturally: `L = c(-1, -2, -3, -4, -5, -6, -7)` would create a
week of lags for daily data.

### Example 3: Multiple Variables with Different Lags

Real forecasting scenarios involve multiple predictors with different lag structures.
You might want:

- 3 lags of the target variable `Y`
- 7 lags of exogenous variable `X1` (a week of history)
- Selected lags of `X2` (perhaps 1, 3, and 7 days)

```{r simple-multi-var}
multi_var <- slice(
    simple_dt_date,
    variables = c("Y", "X1", "X2"),
    walk_on = "date",
    slice_on = "date",
    L = list(
        Y = c(-1, -2, -3),
        X1 = c(-1, -2, -3, -4, -5, -6, -7),
        X2 = c(-1, -3, -7)
    )
)

multi_data <- as.data.table(multi_var)
tail(multi_data, 3)

# Column names:
print(names(multi_data))
```

**Key observations**:

- Each variable can have a different lag structure specified in the list
- `Y` gets 3 lag columns: `Y_1`, `Y_2`, `Y_3`
- `X1` gets 7 lag columns: `X1_1` through `X1_7`
- `X2` gets 3 selected lag columns: `X2_1`, `X2_3`, `X2_7`
- The result is a rich feature matrix with 13 lag features plus target
- All features are properly aligned by date (in the `index` column)

This demonstrates the power of simple slicing: a single `slice()` call creates
a complete autoregressive feature matrix that would require extensive manual
coding otherwise.

### Key Takeaways

**Simple slicing** (`walk_on == slice_on`) is characterized by:

- Creating lag features from the same temporal index
- Progressive complexity: single lag → multiple lags → multiple variables
- List format for `L` parameter allows different lags per variable
- Results are immediately model-ready for autoregressive forecasting
- Proper handling of missing values at temporal boundaries

**IMPORTANT!!**

The suffixes `_1`, `_2`, DO NOT correspond to the actual `L` values directly. They are simply
numbered in along the values provided, existing only for ordering purposes. Take the `L` values for
`X2` in the above example: `c(-1, -3, -7)`. The resulting columns are named
`X2_1`, `X2_2`, and `X2_3`, which should be read as "first `L` value", "second `L` value", and so
on, not "lag 1", "lag 2", etc.


## Keyed Slicing

**Keyed slicing** occurs when you walk on one temporal index but slice on a different
temporal index. This enables integration of data sources with different temporal
resolutions or structures with one caveat: both indexes must be of the same type! This means both
columns must be date-based, datetime-based, or integer-based.

The defining characteristic is: **`walk_on != slice_on`**

Common use cases include:

- Daily forecasts using hourly weather data
- Monthly predictions using daily transaction data
- Integrating forecast/exogenous variables with different granularity

### Concept: Walking and Slicing on Different Indices

When `walk_on` and `slice_on` differ, shapeshiftr:

1. **Walks** through the time points defined by `walk_on`
2. **Slices** from the temporal index defined by `slice_on`
3. **Aligns** the two indices to extract relevant features

This allows you to "reach into" a finer-grained temporal structure while maintaining
a coarser prediction schedule.

### Example 1: Basic Keyed Slicing

Imagine you're forecasting daily outcomes but have access to hourly weather forecasts.
You want to integrate the next 24 hours of forecasted temperature as features for
each daily prediction.

For this example, we'll use a half-hourly dataset and create features for the next 2 hours:

```{r keyed-basic}
data(keyed_dt_datetime)

# Half-hourly forecast data structure:

print(head(keyed_dt_datetime, 15))

keyed_slice <- slice(
    keyed_dt_datetime,
    variables = "Y",
    walk_on = "datetime",
    slice_on = "target_datetime",
    L = 0:4
)

keyed_data <- as.data.table(keyed_slice)

# Keyed slice result (first 2 hours ahead):

print(head(keyed_data, 2))

# Column structure:

print(names(keyed_data))
# 5 half-hourly features total)
```

**Key observations**:

- `walk_on = "datetime"` defines half-hourly time points for predictions
- `slice_on = "target_datetime"` reaches into half-hourly temporal structure
- Positive `L` values create features for future time periods
- `L = 0:4` extracts values at current time (0) and next 4 half-hours
- Result has one row per half-hour with 5 half-hourly features

This is particularly useful when integrating external forecasts or planned
interventions into your forecasting model.

### Understanding the Walk-Slice Relationship

The key to keyed slicing is understanding how `walk_on` and `slice_on` interact:

```
Hourly walk points (walk_on = "date"):

  2025-01-01 01:00      |   2025-01-01 02:00     |   2025-01-01 03:00
                        
           |            |            |           |            |
           v            |            v           |            v

Half-hourly slice points| (slice_on = "datetime")|:

  01:00 01:00 ... 03:00 | 02:00 02:30 ... 04:00  | 03:00 03:30 ... 05:00
  <---- 2 hours ------> | <---- 2 hours ------>  | <---- 2 hours ------>  
```

When you specify `L = list(Y = 0:4)`, shapeshiftr:

1. Starts at each hourly walk point (2025-01-01 01:00, 2025-01-01 02:00, ...)
2. Check what is the resolution of `slice_on` (`target_datetime` is half-hourly)
3. Compute what are the time steps for `L` values (0-4) *in `target_datetime`'s resolution*
   - 0 = current half-hour
   - 1 = next half-hour
   - 2 = two half-hours ahead
   - 3 = three half-hours ahead
   - 4 = four half-hours ahead
4. Finds corresponding datetime in the half-hourly `target_datetime`
5. Creates features `Y_0` through `Y_4`

### Key Takeaways

**Keyed slicing** (`walk_on != slice_on`) is characterized by:

- Integrating data with different temporal resolutions
- Walking on prediction schedule, slicing on data schedule
- Enables incorporation of forecast/exogenous data
- Proper temporal alignment between different granularities

Keyed slicing complements simple slicing: use simple slicing for lag or lead features of
your target and predictors when there is only one temporal index; use keyed slicing to integrate
external data sources with different temporal structures.

## Building Model-Ready Datasets

Now that you understand simple slicing (Section 2) and keyed slicing (Section 3),
let's integrate these concepts into a complete workflow for building model-ready
forecasting datasets.

This section demonstrates the **essential workflow functions**:

- **`merge.slice_artifact()`** - Combine multiple slice results
- **`as.data.table.slice_artifact()`** - Convert to final data.table

These functions transform slice objects into the wide-format data.tables that
modeling functions expect.

### The Complete Workflow

A typical forecasting workflow involves:

1. **Create multiple slices** for different variable groups or lag structures
2. **Merge slice artifacts** to integrate all features
3. **Convert to data.table** for modeling

Let's build a comprehensive forecasting dataset that predicts the day after tomorrow's `Y` (i.e. 
two days ahead) using:

- 3-day lags of target variable `Y` (simple slicing)
- 3-day lags of predictors `X1` and `X2` (simple slicing)
- 2-day leads of forecasted predictors `X1` and `X2` (keyed slicing)
- Categorical features from current day

To avoid confusion between past and future features, we'll make use of the argument `names` of 
`slice()` to clearly label features.

```{r workflow-complete}
data(simple_dt_date)
data(keyed_dt_date)

# ======================================================================
# STEP 1: CREATE LAG FEATURES (Simple Slicing)
# ======================================================================

lags <- slice(
    simple_dt_date,
    variables = c("X1", "X2", "Y"),
    walk_on = "date",
    slice_on = "date",
    L = list(
        X1 = -seq(3),
        X2 = -seq(3),
        Y = -seq(3)
    ),
    names = list(
        X1 = "X1_lag",
        X2 = "X2_lag",
        Y = "Y_lag"
    )
)

# Created a slice_artifact object containing 3-day lags of target variable Y and predictors X1 and X2
# Structure of lags:

print(str(lags, max.level = 1))

# ======================================================================
# STEP 2: CREATE LEAD FEATURES (Keyed Slicing)
# ======================================================================

leads <- slice(
    keyed_dt_date,
    variables = c("X1", "X2"),
    walk_on = "date",
    slice_on = "target_date",
    L = list(
        X1 = seq(2),
        X2 = seq(2)
    ),
    names = list(
        X1 = "X1_lead",
        X2 = "X2_lead"
    )
)

# Created a slice_artifact object containing 2-day leads of predictors X1 and X2
# Structure of lead

print(str(leads, max.level = 1))

# ======================================================================
# STEP 3: MERGE SLICE ARTIFACTS (merge.slice_artifact)
# ======================================================================

# ** This is merge.slice_artifact() **

# Purpose:
#   Combine multiple slice_artifact objects based on temporal alignment.
#   This integrates features from different slicing operations while
#   maintaining proper temporal relationships.

all_predictors <- merge(lags, leads)

# Operation: merge(lags, leads)

# Result:
#   - Contains features from BOTH slice_artifacts
#   - Y lags from lags: Y_lag_1, Y_lag_2, Y_lag_3
#   - X lags from lags: X1_lag_1, X1_lag_2, X1_lag_3, X2_lag_1, X2_lag_2, X2_lag_3
#   - X leads from leads: X1_lead_1, X1_lead_2, X2_lead_1, X2_lead_2
#   - All features temporally aligned by index

# Structure of merged slice_artifact:
print(str(all_predictors, max.level = 2))


# ======================================================================
# STEP 4: CONVERT TO MODEL-READY DATA.TABLE (as.data.table.slice_artifact)
# ======================================================================

# ** This is as.data.table.slice_artifact() **

# Purpose:
#   Convert slice_artifact to wide-format data.table for modeling.
#   This is the final step that produces a dataset ready for lm()
#   glm(), randomForest(), xgboost, or any other modeling function.

model_data <- as.data.table(all_predictors)

# Conversion complete!\n

# Dimensions:", nrow(model_data), "rows ×", ncol(model_data), "columns

# Model-ready dataset (first 6 rows):

print(head(model_data, 6))

# Column structure:
print(names(model_data))

# Summary of key features:
print(summary(model_data[, .SD, .SDcols = -1]))


# ======================================================================
# VERIFICATION: THIS DATASET IS MODEL-READY
# ======================================================================

# ✓ This dataset can be used directly in modeling:

#   # Linear regression example:
#   model <- lm(Y ~ Y_lag_1 + Y_lag_2 + Y_lag_3 + X1_lag_1 + X1_lag_2 + X1_lag_3 +
#                   X2_lag_1 + X2_lag_2 + X2_lag_3, data = model_data)

#   # Or using formula shorthand:
#   model <- lm(Y ~ ., data = model_data[, -'index'])

#   # Generate predictions:
#   predictions <- predict(model, newdata = test_data)

# Key characteristics of model-ready data:
#   - Each row: A prediction task (one time point)
#   - Lag columns: Features (predictors)
#   - Y column: Target variable
#   - index column: Time index for tracking

# Rows with complete cases:

sum(complete.cases(model_data))

print(nrow(model_data))
```

### Workflow Summary

The three-step workflow is:

```r
slice() → merge() → as.data.table()
```

**Step 1: `slice()`** - Create temporal features

- Use simple slicing for lag features (`walk_on == slice_on`)
- Use keyed slicing for external data (`walk_on != slice_on`)
- Create separate slices for different variable groups

**Step 2: `merge.slice_artifact()`** - Integrate multiple slices

- Combines slice_artifact objects with temporal alignment
- Essential for integrating different feature groups
- Maintains proper temporal relationships across all features

**Step 3: `as.data.table.slice_artifact()`** - Final conversion

- Transforms slice_artifact to wide-format data.table
- Produces immediately model-ready dataset
- Standard tabular format works with any modeling framework

### Key Takeaways

**Building model-ready datasets** requires understanding:

- **`slice()`** creates temporal features (covered in Sections 2-3)
- **`merge.slice_artifact()`** integrates multiple slice results
- **`as.data.table.slice_artifact()`** produces final modeling dataset

**Workflow patterns**:

- Create multiple slices for different variable groups
- Merge slices to combine all features
- Convert to data.table as final step
- Result is wide-format, model-ready data

**Best practices**:

- Keep related variables in the same slice when possible
- Use `merge()` to combine slices with proper temporal alignment
- Always verify temporal alignment after merging
- Check for complete cases before modeling

This workflow forms the foundation for all temporal feature engineering in
shapeshiftr. Sections 5-8 (future sprints) will cover advanced patterns,
missing value strategies, and production deployment.
